---
title: "Analyzing *Syngnathus floridae* RNAseq Data from MSU"
author: "Coley Tosto"
date: "`r Sys.Date()`"
output: 
    html_document:
        code_folding: show
        toc: yes
        toc_float: yes
        number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

``` {r library, include = FALSE}

```

``` {r functions}
```

``` {r read-data}

```

# Pre-assembly QUality Control and Filtering

## Trimming the Reads with Trimmomatic
[Trimmomatic](http://www.usadellab.org/cms/index.php?page=trimmomatic) is commonly used for Illumina paired-end and single ended data. There are a lot of different trimming steps that can be performed and parameters that correspond to each step. Currently for version **0.40** of trimmomatic these are the different steps:\
\

1.  **ILLUMINACLIP**: Cut adapter and other illumina-specific seuqences from the read.\
    `:<fastaWithAdaptersEtc>:<seed mismatches>:<palindrome clip threshold>:<simple clip threshold>`\
    -   fastaWithAdaptersEtc = specifies the path to a fasta file containing all the adapters, PCR sequences etc.
    -   seedMismatches = the max. mismatch count which will still allow a full match to be performed.
    -   palindromeClipThreshold = how accurate the match between the two 'adapter ligated' reads must be for PE palindrome read alignment
    -   simpleClipThreshold = how accurate the match between any adapterect. sequence must be against a read.
2.  **SLIDINGWINDOW**: Perform a sliding window trimming, cutting once the average quality within the window falls below a designated threshold.\
    `:<windowSize>:<requiredQuality>`\
    -   windowSize = the number of bases to average across
    -   requiredQuality = the average quality required
3.  **LEADING**: Cut bases off the start of a read, if below a designated threshold quality.\
    `:<quality>`\
    -   quality = the minimum quality required to keep a base
4.  **TRAILING**: Cut bases off the end of a read, if below a designated threshold quality.\
    `:<quality>`\
    -   quality = the minimum quality required to keep a base
5.  **CROP**: Cut the read to a specified length.\
    `:<length>` 
    -   length = the number of bases to keep, from the start of the read
6.  **HEADCROP**: Cut the specified number of bases from the start of a read.\
    `:<length>`\
    -   length = the number of bases to remove from the start of the read
7.  **MINLEN**: Drop the read if it is below a specified length.\
    `:<length>`
    -   length = the minimum length of reads to be kept

For **paired-end** data, two input files are specified and 4 output files. Two for the **paired** output where both of the reads survived the processing, and two for the corresponding **unpaired** output where one read survived but it's partner did not.\
\
**Trimming occurs in the order which the steps are specified on the command line**. Recommended that if the adapter clipping is required it is done as early as possible.

### Installing Trimmomatic

Trimmomatic was installed on the RCC through conda in a conda environment called **Trim**.

```
conda create -n Trim
conda activate Trim
conda install trimmomatic
```

### Running trimmomatic across Multiple Pairs of Reads

```{bash run_trimmomatic, eval=FALSE}
#!/bin/bash

data=/home/rccuser/20220902_mRNASeq_PE150/ ##This is the location of the raw reads

for fq in ${data}*_R1.fastq.gz
        do
        base=$(basename $fq _R1.fastq.gz)
        echo "Running trimmomatic for ${base}..."
        time trimmomatic PE -threads 16 $fq ${data}${base}_R2.fastq.gz \
                ${data}trimmed/${base}_paired_R1.fastq.gz ${data}trimmed/${base}_unpaired_R1.fastq.gz \    
                ${data}trimmed/${base}_paired_R2.fastq.gz ${data}trimmed/${base}_unpaired_R2.fastq.gz \    
                ILLUMINACLIP:NexteraPE-PE.fa:2:30:10 HEADCROP:12 LEADING:3 TRAILING:10 SLIDINGWINDOW:4:15 MINLEN:50
done
```

-   Before running the script, change the `trimmomatic` line to `echo "..."` to make sure all of the variables are working correctly.

-   Then remove the `echo ""` and run the script as `nohup bash trim_script.sh > trim.out 2>&1 &`.\

-   The **NexteraPE-PE.fa** file was pulled from the [trimmomatic github](https://github.com/usadellab/Trimmomatic/tree/main/adapters) using `wget https://github.com/usadellab/Trimmomatic/blob/main/adapters/NexteraPE-PE.fa`


## Using Kraken2 to remove biological contamination
### Create a new environment and install Kraken2 in it

```
conda create -n kraken2
conda activate kraken2
conda install kraken2
```

### Creating the Standard Kraken2 Database {.tabset}
The Kraken2 Database can be generated in one of two ways

#### Builidng it by Hand {.unnumbered}
Kraken2 has a built in function `kraken2-build` that allows you to build whatever database you are wanting. For example if you want to generate the standard kraken2 database (That includes archea and bacteria):

```
mkdir kraken2_standard_db

kraken2-buid --standard --db kraken2_standard_db/
```

If you run into the following error:
"rsync_from_ncbi.pl: unexpected FTP path (new server?) for https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/900/128/725/GCF_900128725.1_BCifornacula_v1.0" 

You will need to change some of the code in the file `rsync_from_ncbi.pl`. If Kraken 2 was installed via conda the file will be located in `/anaconda3/envs/kraken2/share/kraken2-2.1.2-4/libexec`. Use `nano` to edit the file and on line **46** replace _ftp_ with _((ht|f)tp(s?))_ so the whole line for 46 would look like this:

`if (! ($full_path =~ s#^((ht|f)tp(s?))://${qm_server}${qm_server_path}/##)) {`


#### Uploading a pre-built database {.unnumbered}
This [website](https://benlangmead.github.io/aws-indexes/k2) also has all of the different datasets pre-built and available for download. To access one of these pre-built databases through the RCC directly use:

```
wget https://genome-idx.s3.amazonaws.com/kraken/k2_pluspfp_20230314.tar.gz #Using the link associated to the dataset of choice

mkdir kraken2_pluspfp_database
tar -xvf k2_pluspfp_20230314.tar.gz -C kraken2_pluspfp_database/
```

### Creating a for loop to run multiple samples at a time
```{bash run_kraken2, eval = FALSE}
#!/bin/bash

#Create arguments
ref_fasta=$1 #desired reference database
input_dir=$2 #Input directory with the location on the reads
output_dir=$3 #Desired location for the output

## Loop through all pairs of reads in the input directory
for pair in $2/*_R1.fq.gz
        do

        #Extract the sample name from the file name
        sample=$(basename $pair _R1.fq.gz)

        ##Echo the sample name it is currently running
        echo "Running Kraken2 for ${sample}..."

        #Define the paths to the input and output files for this sample
        input1=$2/${sample}_R1.fq.gz
        input2=$2/${sample}_R2.fq.gz

        #Run Kraken2 on this pair of reads
        time kraken2 --threads 16 --db $1 --paired $input1 $input2 --unclassified-out $3/${sample}#.fq --report $3/${sample}.log

done
```

This script was run as `nohup bash bash_scripts/kraken2.sh ../kraken2_pluspfp/ > kraken2.log 2>&1 &`

## Using SortMeRNA to remove rRNA contamination
### Create a new environment and install SortMeRNA in it

```
conda create -n sortmerna
conda activate sortmerna
conda install sortmerna

which sortmerna #check location of installation
  /home/rccuser/anaconda3/envs/contamination/bin/sortmerna
```

### Test the installation of SortMeRNA

```
sortmerna --version
    SortMeRNA version 4.3.6
    Build Date: Aug 16 2022
    sortmerna_build_git_sha:@db8c1983765f61986b46ee686734749eda235dcc@
    sortmerna_build_git_date:@2022/08/16 11:42:59@
```

### Download the latest SortMeRNA databases (v4.3.4)
The `database.tar.gz` file can also be downloaded from [this link](https://github.com/sortmerna/sortmerna/releases).

```
wget https://github.com/biocore/sortmerna/releases/download/v4.3.4/database.tar.gz
mkdir rRNA_databases_v4.3.4
tar -xvf database.tar.gz -C rRNA_databases_v4.3.4
```

These .fasta files are what will be used in the next step for our reference file. There are 4 fasta files associated with `v4.3.4`.

  1. `smr_v4.3_default_db.fasta`: this FASTA file contains the **default** SortMeRNA database, which is a comprehensive rRNA database derived from the SILVA and Rfam databases. This database includes Bacteria 16S/23S (SILVA), Archaea 16S (SILVA), Eukarya 18S/28S (SILVA),5S/5.8s rRNA (Rfam).
  
  2. `smr_v4.3_sensitive_db_rfam_seeds.fasta`: this FASTA file contatins **seed sequences** of the covariance models (CMs) from the Rfam database. Rfam is a comprehensive collection of RNA families, each represented by a CM that describes the consensus secondary structure of the RNA family. These CMs are used to annotate RNA sequences in genomes and metagenomes. The sensitivity of SortMeRNA can be increased by including this database in the analysis, but it can also increase the runtime and memory requirements of the analysis.
  
  3. `smr_v4.3_fast_db.fasta`: this FASTA file contains **a subset of sequences** from the default SortMeRNA database. Here, the number of sequences in each database is reduced to improve the speed of the analysis. This file would be intended for use in cases where speed is a priority and a lower sensitivity is acceptable.
  
  4. `smr_v4.3_sensitive_db.fasta`: this FASTA file contains **a more comprehensive set of sequences** than the default SortMeRNA database. The sensitive database is intended for use in cases where higher sensitivity is required, such as in metagenomic or environmental sequencing studies, but similar to the second FASTA file it will also increase the runtime and memory requirements of the analysis.
  
**More than one of theses reference databases can be used at a time if desired**. To do this you add `--ref` twice in the script (e.g. `--ref smr_v4.3_default_db.fasta --ref smr_v4.3_sensitive_db_rfam_seeds.fasta` ).


### Creating a for loop to run multiple samples at a time
```{bash run_sortmeRNA, eval = FALSE}
#!/bin/bash

#Create arguments
input_dir=$1 #location of the reads
ref_fasta=$2 #desired reference fasta
output_dir_rrna=$3 #desired location for the reads that are rRNA
output_dir_norrna=$4 #desired location for the reads that are NOT rRNA

## Loop through all pairs of reads in the input directory
for pair in $1/*_R1.fastq.gz
        do

        #Extract the sample name from the file name
        sample=$(basename $pair _R1.fastq.gz)

        #Extract Fish ID
        ID=$(basename $pair _paired_R1.fastq.gz)

        ##Echo the sample name it is currently running
        echo "Running SortMeRNA for ${ID}..."

        #Define the paths to the input and output files for this sample
        input1=$1/${sample}_R1.fastq.gz
        input2=$1/${sample}_R2.fastq.gz

        #Run SortMeRNA on this pair of reads
        time sortmerna --threads 16 --ref $2 --reads $input1 --reads $input2 --fastx --aligned $3/${ID} --other $4/${ID} --out2

        #Remove SortMeRNA intermediate files before running again
        rm -r /home/rccuser/sortmerna/run/kvdb
done
```

This script was run as `nohup bash bash_scripts/sortmerna.sh floridae_trimmed ../rRNA_databases_v4.3.4/smr_v4.3_fast_db.fasta floridae_rrna floridae_norrna > sortmerna.log 2>&1 &`

## Doing a k-mer based correction with RCorrector
Rcorrector (RNA-seq error CORRECTOR) is a kmer-based error correction method that is used on RNA-seq data. This program allows for a correction of random sequencing errors in Illumina RNA-seq reads. The downside is you may lose information on things like SNPs and other sequence variants.

### Install Rcorrector 

1. Clone the GitHub Repo: `git clone https://github.com/mourisl/rcorrector.git`

2. Run `make` in the repo directory.

### Creating a for loop to run multiple samples at a time
```{bash run_Rcorrector, eval = FALSE}
#!/bin/bash

#Create arguments
rcorrector_path=$1 #Path to the run_rcorrector.pl file
input_dir=$2 #Path to the input directory containing the reads
output_dir=$2 #Path to the desired output location

##Loop through all pairs of reads in the directory
for pair in $2/*_R1.fq
        do

        #Extract sample name from the file name
        sample=$(basename $pair _R1.fq)

        #Echo the sample name that is currently running
        echo "Running rcorrector for ${sample} ..."

        #Run rcorrector on this pair of reads
        time perl $1/run_rcorrector.pl -t 16 -1 $2/${sample}_R1.fq -2 $2/${sample}_R2.fq -od $3

done
```

This script was run as `nohup bash bash_scripts/rcor.sh ../../rcorrector floridae_nobio floridae_kmer_corrected/ > rcor.log 2>&1 &`

# De novo transcriptome assembly

## Installing Trinity
Trinity requires the installation of several programs, to get around this Trinity was installed via a Docker in the RCC and must be used within one. **The first time that you use Trinity you must pull the latest Docker image for Trinity** like so:

```
docker pull trinityrnaseq/trinityrnaseq

```

## Running Trinity 
Once started, a successful Trinity run will take days to complete. There are a lot of extra components you can add onto a Trinity run (View the [Trinity User Manual](https://github.com/trinityrnaseq/trinityrnaseq/wiki/Running-Trinity) for more information) but here are the main components you should specify:

1.  **SeqType**: Signifies the type of reads (fa or fq).
    `--seqType <string>`

2.  **Max Memory**: Suggested max memory to use by Trinity where limiting can be enabled.
    `--max_memory <string>`
    -   This should be based on the available memory for your computer. For example `--max_memory 20G`
    
3.  **Read Locations**: Provide exact file locations for the left and right reads. One or more file names separated by commas, NOT spaces.
    `--left <string>`
    `--right <string>` 
    -   The left and right reads must be paired up sequentially, if XXX32_R1 is mentioned first for the left, XXX32_R2 must also be first for the right.
    -   If running across multiple samples `--samples_file <string>` can be used rather than typing out the location for each individual. See below for more details.

4.  **CPU**: Tell Trinity the number of CPUs to use, default is 2.
    `--CPU <int>`
    -   This should be based on the number of cores the computer has and it must be an even number.
    
5.  **Output Directory**: Name of the directory for output.This will be created if it doesn't already exist.
    `--output <string>`
    -   Good to include trinity in the name as a safety precaution! Don't forget to include the full file path.
    
### Using multiple reads
#### Creating a Sample Name File
There are two options when using multiple files to generate your de novo transcriptome assembly. Each sample can be typed out individually, ensuring that the two reads are paired up by order. Alternatively, the argument `--samples_file` can be used instead. With this argument you must provide a tab-delimited text file indicating biological replicate relationships. This is the example provided on the [Trinity Website](https://github.com/trinityrnaseq/trinityrnaseq/wiki/Running-Trinity):

         cond_A    cond_A_rep1    A_rep1_left.fq    A_rep1_right.fq
         cond_A    cond_A_rep2    A_rep2_left.fq    A_rep2_right.fq
         cond_B    cond_B_rep1    B_rep1_left.fq    B_rep1_right.fq
         cond_B    cond_B_rep2    B_rep2_left.fq    B_rep2_right.fq
         
The samples file that was provided for the script below was adapted into this format:

```
FLG2F   FLG2F7  /home/rccuser/shared/coley_files/floridae_kmer_corrected/FLG2F7_R1.fq.gz        /home/rccuser/shared/coley_files/floridae_kmer_corrected/FLG2F7_R2.fq.gz
FLG3F   FLG3F1  /home/rccuser/shared/coley_files/floridae_kmer_corrected/FLG3F1_R1.fq.gz        /home/rccuser/shared/coley_files/floridae_kmer_corrected/FLG3F1_R2.fq.gz
FLG3F   FLG3F2  /home/rccuser/shared/coley_files/floridae_kmer_corrected/FLG3F2_R1.fq.gz        /home/rccuser/shared/coley_files/floridae_kmer_corrected/FLG3F2_R2.fq.gz
FLG3M   FLG3M5  /home/rccuser/shared/coley_files/floridae_kmer_corrected/FLG3M5_R1.fq.gz        /home/rccuser/shared/coley_files/floridae_kmer_corrected/FLG3M5_R2.fq.gz
FLG3M   FLG3M7  /home/rccuser/shared/coley_files/floridae_kmer_corrected/FLG3M7_R1.fq.gz        /home/rccuser/shared/coley_files/floridae_kmer_corrected/FLG3M7_R2.fq.gz
...
```
The last two columns contain the **full** location to the desired reads.

#### Script
```{bash trinity_script, eval = FALSE}
#!/bin/bash

#Create arguments
samples_file=$1 #File containing sample names and locations
out_dir_name=$2 #Desired name for the output directory

sudo docker run --rm -v`pwd`:`pwd` trinityrnaseq/trinityrnaseq Trinity --seqType fq --max_memory 188G \
        --samples_file $1 \
        --normalize_by_read_set \
        --CPU 16 \
        --output `pwd`/$2
```

# Post-Assembly Quality Control
## The Bellerophon Pipeline
This pipeline was created to remove falsely assembled chimeric transcripts in de novo transcriptome assemblies. It integrates several tools including **TransRate**, **CD-HIT**, and **BUSCO**.

### Installing the pipeline
```
git clone https://github.com/JesseKerkvliet/Bellerophon.git

```
There were issues trying to install the pipeline via Conda. Due to incompatibilities between packages. Because of this I am instead trying to install the pipeline with Vagrant VM. First Vagrant and virtualbox must be installed.

```
sudo apt install vagrant
vagrant --version
  Vagrant 2.2.19

sudo apt install virtualbox
virtualbox -h
  Oracle VM VirtualBox VM Selector v6.1.38_Ubuntu
  
##Create the Vagrant File
vagrant init bellerophon/bellerophon --box-version 1.0

##Start the Virtual Machine
vagrant up
```
## Using BUSCO to assess composition or 'completeness' of assembly
[BUSCO](https://busco.ezlab.org/busco_userguide.html) (Benchmarking Universal Single-Copy Orthologs) maintains curated sets of universal single-copy genes from OrthoDB and assess the completeness if an assembly by how many of the **universal** genes have matches in the input data and whether those matches are duplicated, fragmented, or full length. 

The base command to run a BUSCO assessment looks something like this:
    
    busco -i <SEQUENCE_FILE> -l <LINEAGE> -o <OUTPUT_NAME> -m <MODE>
    
      -i is a path to your FASTA file, either nucleotide or protein fast file
      -l is the lineage, the manually picked dataset to be used for the assessment
      -m is which analysis mode to run, wither genome (geno), protein (prot), or transcriptome (tran)
      -o is the name of the output directory

Use `busco --list-datasets` to print the full list of clades that can be used for the lineage.

### Installing BUSCO from Bioconda

```
conda update -n base conda

conda create -n busco -c conda_forge -c bioconda busco=5.2.2
conda activate busco

#Double check version
busco -v
```

### Running BUSCO on a single transcriptome
The lineage for _S. floridae_ was chosen as `actinopterygii_odb10` since it is the closest clade provided with BUSCO.

```{bash run_busco, eval = FALSE}
#!/bin/bash

#Create arguments
transcriptome=$1 #Output fasta file from Trinity
lineage=$2 #chosen dataset for assessment
out_dir_name=$3 #Desired name for the output directory

busco -i $1 -l $2 -m tran -o $3 -c 16


```
This script was run as `nohup bash bash_scripts/busco_tran.sh trinity_out_dir_floridae_May2023.Trinity.fasta actinopterygii_odb10 busco_out_floridae > busco.log 2>&1 &`

### Interpreting BUSCO results

# Alignment and Abundance Estimations
## Using Salmon
Salmon was installed in a conda environment
```
conda create -n salmon
conda activate salmon
conda install samlon
```
### Generating the index
The function `index` was then used to generate an index on the transcriptome. This index is the structure that salmon uses to quasi-map RNA-seq reads during the quantification step. **This index has to be created only once per transcriptome**

```{bash salmon-index, eval = FALSE}
#!/bin/bash

#Create the arguments
transcriptome_file=$1 #This is the output .fasta file from Trinity
output_index_name=$2
desired_kmer_length=$3

salmon index -t $1 -i $2 -k $3

```

The above script was run as: `nohup bash bash_scripts/salmon_txome_indicies.sh trinity_out_dir_floridae_May2023.Trinity.fasta floridae_salmon_index 31 > salmon.log 2>&1 &`

A k-mer length of 31 was chosen as a recommendation from the [salmon documentation](https://salmon.readthedocs.io/en/latest/salmon.html#using-salmon). If I end up seeing a smaller mapping rate than expected I will generate another index using a smaller k-mer length.

### Quantifying the samples
Once the index is built, the samples can be quantified.

```{bash salmon-quant, eval = FALSE}
#!/bin/bash

#Create arguments
input_dir=$1
index_file=$2
output_dir=$3

for fq in $1/*_R1.fq.gz
        do

        #Extract sample name from the file
        sample=$(basename $fq _R1.fq.gz)

        #Echo sample name that is currently running
        echo "Processing sample ${sample} ..."

        #Quantify this pair of reads
        salmon quant -i $2 -l A \
                -1 $1/${sample}_R1.fq.gz \
                -2 $1/${sample}_R2.fq.gz \
                -p 16 --validateMappings -o $3/${sample}_quant

done
```

This script was run as: `nohup bash bash_scripts/salmon_quant.sh floridae_kmer_corrected floridae_salmon_index/ floridae_salmon_quant > salmon_quant.log 2>&1 &`
