---
title: "Differential Expression Analysis in _Stigmatopora nigra_"
output:
  bookdown::pdf_document2:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    toc: true
header-includes: >
  \usepackage{lipsum}
  \usepackage{float}
  \floatplacement{figure}{H}
editor_options:
  chunk_output_type: console
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir='../',fig_path="../imgs/")
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```

``` {r library, include = TRUE, message = FALSE, warning = FALSE}
#This is a cohesive list of all the libraries used in this document
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
library(PCAtools)
library(ggpubr)
library(cowplot)
library(UpSetR)
library(ggplot2)
library(kableExtra)
library(tidyverse)
library(knitr)
library(magick)
```
``` {r read-data}
#The abundance matrix generated via salmon and tximport to be used for the DE analysis
txi.nigra <- readRDS("txi.salmon_SN.RDS")

#The samples file generated for tximport
samples <- read.table("SN_samples.txt", header = TRUE)

#Make sure the conditions are in the samples file as a factor
samples$sex <- as.factor(samples$sex)
samples$organ <- as.factor(samples$organ)

```

The package `DESeq2` was used for the differential expression analysis outlined below.

# Single factor analysis - Comparing Males v Females across all organs
To analyze your data with DESeq2 you must first generate the DESeqDataSet. In order to do this we need the abundance matrix generated with `tximport` and a `samples` file that lays out all of the conditions. The model for this single factor analysis was run as counts ~ Sex.

```{r DESeqDataSet, message=FALSE, warning=FALSE}
#Create the DESeq dataset
dds_SN <- DESeqDataSetFromTximport(txi.nigra, 
                                   colData = samples,
                                   design = ~ sex)

```

The data is then pre-filtered to remove low gene counts before running further DESeq2 functions. By doing this we remove rows in which there are very few reads thus reducing the memory size of the `dds` object and increasing the speed at which we can use the transformation and testing functions in DESeq2.

The cutoff here was to remove rows that had counts fewer than 10 across all samples.

```{r pre-filtering, message=FALSE, warning=FALSE}
#only keeping rows that have at lead 10 reads total
keep <- rowSums(counts(dds_SN)) >= 10
dds_SN <- dds_SN[keep, ]

```

After filtering we can now perform the standard differential expression analysis that is wrapped into DESeq2.

```{r diff-exp, message=FALSE, warning=FALSE}
#Generate the expression values
dds_SN_exp <- DESeq(dds_SN)

#Compile the results
res <- results(dds_SN_exp)
res
```
