---
title: "Differential Expression Analysis in _Stigmatopora nigra_"
output:
  bookdown::pdf_document2:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    toc: true
header-includes: >
  \usepackage{lipsum}
  \usepackage{float}
  \floatplacement{figure}{H}
editor_options:
  chunk_output_type: console
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir='../',fig_path="../imgs/")
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```

``` {r library, include = TRUE, message = FALSE, warning = FALSE}
#setwd("~/Documents/GitHub/pipefish_sbge/stigmatopora/S_nigra")
#This is a cohesive list of all the libraries used in this document
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
library(PCAtools)
library(ggpubr)
library(cowplot)
library(UpSetR)
library(ggplot2)
library(kableExtra)
library(tidyverse)
library(knitr)
library(magick)
library(multcompView)
```
``` {r read-data}
#The abundance matrix generated via salmon and tximport to be used for the DE analysis
txi.nigra <- readRDS("data/txi.salmon_SN.RDS")

#The samples file generated for tximport
samples <- read.table("data/SN_samples.txt", header = TRUE)

#Make sure the conditions are in the samples file as a factor
samples$sex <- as.factor(samples$sex)
samples$organ <- as.factor(samples$organ)

```

The package `DESeq2` was used for the differential expression analysis outlined below.

# Single factor analysis - Comparing Males v Females across all organs
To analyze your data with DESeq2 you must first generate the DESeqDataSet. In order to do this we need the abundance matrix generated with `tximport` and a `samples` file that lays out all of the conditions. The model for this single factor analysis was run as counts ~ Sex.

```{r DESeqDataSet, message=FALSE, warning=FALSE}
#Create the DESeq dataset
dds_SN <- DESeqDataSetFromTximport(txi.nigra, 
                                   colData = samples,
                                   design = ~ sex)

```

The data is then pre-filtered to remove low gene counts before running further DESeq2 functions. By doing this we remove rows in which there are very few reads thus reducing the memory size of the `dds` object and increasing the speed at which we can use the transformation and testing functions in DESeq2.

The cutoff here was to remove rows that had counts fewer than 10 across all samples.

```{r pre-filtering, message=FALSE, warning=FALSE}
#only keeping rows that have at lead 10 reads total
keep <- rowSums(counts(dds_SN)) >= 10
dds_SN <- dds_SN[keep, ]

```

After filtering we can now perform the standard differential expression analysis that is wrapped into DESeq2.

```{r diff-exp, message=FALSE, warning=FALSE}
#Generate the expression values
dds_SN_exp <- DESeq(dds_SN)

#Compile the results
res <- results(dds_SN_exp)
res
```

Once that has finished we can now start exploring some of the single-factor analysis results
```{r investigate-DESeq-results-SF}
##Ordering our results based on p-value
resOrdered <- res[order(res$pvalue),]
resOrdered

summary(res)

#How many ADJUSTED p-values were less than 0.1?
sum(res$padj < 0.1, na.rm = TRUE)

#Looking at  an alpha=0.05, the default is 0.1
res05 <- results(dds_SN_exp, alpha = 0.05)
summary(res05)
sum(res05$padj < 0.05, na.rm = TRUE)

```

## Visualizing the results
### MA-plot - MvF Differntial expression
Generate an MA-plot to show the log2 fold changes attributable to sex over the mean of normalized counts for all of the samples in the `dds`. Points will be colored if the adjusted p-value is less that 0.1.

```{r MA-plot, echo=FALSE, fig.cap="LogFC versus the mean normalized count for all of the genes. Points that are blue have a p-value less than 0.1. The two plots are the same with the only difference coming from an adjusted y-limit. Points that are triangles represent genes with a fold change higher/lower than the y-limit."}

par(mfrow=c(1,2))
plotMA(res, ylim = c(-2,2))
plotMA(res)

```

### Heatmap - overall expression
We can also generate a heat map to look ar overall expression levels across our samples. Note, this is not differentially expressed genes.

```{r heatmap, echo=FALSE, fig.cap= "Heatmap showing the expression level across all organs for the top 20 genes with the highest expression levels."}
#Pull out the top 20 rows with the most expression
select <- order(rowMeans(counts(dds_SN_exp, normalized = TRUE)),
                decreasing = TRUE)[1:20]
df <- as.data.frame(colData(dds_SN)[,c("sex", "organ")])

#Transform the data
vsd <- vst(dds_SN_exp, blind=FALSE)

#Run the heat map with the function pheatmap
pheatmap(assay(vsd)[select,], 
         cluster_rows = FALSE, 
         show_rownames = TRUE, 
         cluster_cols = FALSE, 
         annotation_col = df)
```

From the heatmap (Fig. \@ref(fig:heatmap)) we can see that for all of these top 20 expressed genes the highest expression in found in the liver. There are also a few genes that appear to be highly expressed across all of the tissue types. With this heatmap we can also pull out the names of the Trinity genes that are showing this high expression and BLAST the corresponding sequences to see what the genes are.

```{r heatmap_genes}
#Pull out the corresponding trinity_geneIDS that are plotted in the heatmap
heatmap_TG <- cbind(row.names(assay(vsd)[select,]))
#Run once to create the file and then commented out
#write.table(heatmap_TG,
            #'heatmap_trinitygenes.txt',
            #sep = "",
            #quote=FALSE,
            #row.names = FALSE,
            #col.names = FALSE)
```

### Sample-dist Heatmap

We can then generate a sample-to-sample distances heatmap that will give an overview of the similarities and differences between our samples.

```{r sample-dist, echo=FALSE, fig.cap="Sample-to-sample distances heatmap showing the similarities and differences between all of the samples. The darker the color, the more similar the samples are. The diagonal line of very dark blue represents the comparissons between the same samples.", warning=FALSE}
#Calculate the sample-to-sample distances from the transformed count matrix
sampleDists <- dist(t(assay(vsd)))

#Transform into a matrix
sampleDistsMatrix <- as.matrix(sampleDists)
rownames(sampleDistsMatrix) <- paste(vsd$sex, vsd$organ, sep = "-")
colnames(sampleDistsMatrix) <- vsd$ID

#Run the heatmap
colors <- colorRampPalette(rev(brewer.pal(9, 'Blues')))(255)
pheatmap(sampleDistsMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists, 
         col = colors)
```

### PCA plots
Next we can look at PCA plots for our samples to see how our different samples are grouping. The size of the points for these plots will be scaled by the mating success of the individual to see if that may be at all correlated with the sample grouping.

```{r pca-scaled-organ, echo=FALSE, fig.cap="Principal components analysis reflects that most variation corresponds to differences in expression between organs (green: gill; yellow: gonad; pink: liver), rather than variation due to sex (circle = female; triangle = male) or mating status (the size of points reflects an individual’s mating success). The first axis explains 32% of variation in the dataset, with gills and liver, two types of somatic tissues, sitting on opposite ends of that axis. The second axis explains 23% of the variation in gene expression, with somatic tissues on the opposite side of the axis from the gonads.", warning=FALSE}

#Generate the PCA dataframe
pca <- prcomp(t(assay(vsd)))
pca_plotting_data<-as.data.frame(pca$x)

#Add the metadata to the PCA dataset
pca_plotting_data$organ <- samples$organ
pca_plotting_data$sex <- samples$sex

#Calculate the percent variation attributed to each axis 
pcaData <- plotPCA(vsd, intgroup = c("organ", "sex"), returnData = TRUE, ntop = nrow(dds_SN))
percentVar <- round(100 * attr(pcaData, "percentVar"))


my_colors <- c("Gill" = "#6E8B3D75", "Gonad" = "#EEB42275", "Liver" = "#EE826275")
my_shapes <- c(16, 17)

par(mfrow=c(1,2))
ggplot(pca_plotting_data, aes(x=PC1, y=PC2, color = organ, shape = sex)) +
  scale_color_manual(values = my_colors) +
  scale_shape_manual(values = my_shapes) +
  geom_point(aes(size = organ)) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"))
```

```{r pca-scaled-sex, echo=FALSE, fig.cap="The same Principal Components Analysis as above but with color now representing sex and shape representing organ type (left) as well as the comparisson of PC1 and PC3 which shows variation in expression patterns between males and females, particularly in the gonads, explains approximately 10% of the variance in gene expression data, as reflected by the third principal components axis."}

par(mfrow=c(1,2))
my_colors2 <- c("F" = "#7fc97f75", "M" = "#beaed475" )
my_shapes2 <- c(16, 17, 15)
pc1v2 <- ggplot(pca_plotting_data, aes(x=PC1, y=PC2, color = sex, shape = organ)) +
  scale_color_manual(values = my_colors2) +
  scale_shape_manual(values = my_shapes2) +
  geom_point(aes(size = organ)) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"))

pc1v2

pc1v3 <- ggplot(pca_plotting_data, aes(x=PC1, y=PC3, color = sex, shape = organ)) +
  scale_color_manual(values = my_colors2) +
  scale_shape_manual(values = my_shapes2) +
  geom_point(aes(size = organ)) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC3: 10% variance")) + 
  coord_fixed() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"))

pc1v3

```

# Multifactor design - Comparing M v F across the diff tissue types
If we investigate the column data of our DESeq dataset we can see that each sample has both a sex and organ type attached to it, we will be using these two factors in our multi-factor analysis. In this multi-factor analysis the model was run as counts ~ group, where group included both the sex and the organ type (i.e. MLiver, FLiver, etc.). The sample "S34" appeared to a bit of an outlier in the above analysis, particularly clear in the sample-dist heatmap and the PCA, so it will be removed from the analysis going forward.

```{r multi-factor-analysis, message=FALSE, warning=FALSE}
#Create an additional column the has the two conditions combined(sex and organ type)
samples$group <- factor(paste0(samples$sex, samples$organ))

##Create a copy of the DESeq dataset
ddsMF_SN <- DESeqDataSetFromTximport(txi.nigra,
                                     colData = samples,
                                     design = ~ group)

##Remove S34 from the dataset
ddsMF_SN <- ddsMF_SN[, ddsMF_SN$ID != "S34"]

##Filter the dataset, only keeping rows that have at least 10 reads total
keep <- rowSums(counts(ddsMF_SN)) >= 10 #& rowSums(counts(ddsMF_SN)) < 1e6
ddsMF_SN <- ddsMF_SN[keep, ]

#Run the differential expression analysis
ddsMF_SN_exp <- DESeq(ddsMF_SN)
resultsNames(ddsMF_SN_exp)
```

## Invesitgate the results of the differential expression
Thanks to the multi-factor analysis we can now explore differential expression between all of the different combinations:

  - Male Liver v. Female Liver
  - Male Gill v. Female Gill
  - Male Gonad v. Female Gonad
  - All of the within sex tissue comparisons (e.g. Male Liver v. Male Gill, etc.)

### M-F Liver Comparisson
```{r m-f-liver-results}
##Pulling out the liver M-F results with an alpha of 0.05
liver_con_res <- results(ddsMF_SN_exp, contrast = c("group", "MLiver", "FLiver"),
                         alpha = 0.05)
liver_con_res$trin_geneid <- rownames(liver_con_res)
head(liver_con_res)

summary(liver_con_res)
```

There are many criteria that have been employed previously to denote sex bias. For this study we are classifying sex-biased genes as **genes with a p-value < 0.05 AND a logFC $\ge$ |2|**. With that criteria in the liver there are `r sum(liver_con_res$padj <= 0.05 & liver_con_res$log2FoldChange >= 2, na.rm = TRUE)` male-biased genes and `r sum(liver_con_res$padj <= 0.05 & liver_con_res$log2FoldChange <= -2, na.rm = TRUE)` female-biased genes. 

I have then pulled out all of the male-biased, female-biased, and non-biased genes based on the criteria outlined above. I determined non-biased genes as a p-vaule > 0.05.

```{r liver-sex-biased}
#Removing the rows where padj. is NA in results
liver_con_res_noNA <- liver_con_res[!is.na(liver_con_res$padj),]
summary(liver_con_res_noNA) #We can now see that there are no outliers or low counts since the NAs have been removed

#Creating a vector that contains all of the male-biased and female-biased genes in the liver
liver_mal_biased <- liver_con_res_noNA[which(liver_con_res_noNA$log2FoldChange >= 2
                                             & liver_con_res_noNA$padj <= 0.05),]
liver_fem_biased <- liver_con_res_noNA[which(liver_con_res_noNA$log2FoldChange <= -2 
                                             & liver_con_res_noNA$padj <= 0.05),]

#Creating an object that contains all of the non-biased genes in the liver
liver_non_biased <- liver_con_res_noNA[which(liver_con_res_noNA$padj > 0.05),]

```

I will be generating a table that outlines the genes that are the most male or female biased in each organ type. To do this I need to pull the top 50 differentially expressed genes in males or females, get the Trinity gene IDs and then BLAST the corresponding sequences. Here I am pulling out the Trinity gene IDs for those genes.

```{r top50-liver}
#Creating a subset of the results where the p-value is less than 0.05
liver_con_res_p05 <- liver_con_res_noNA[liver_con_res_noNA$padj <= 0.05, ]


#Pulling the top 50 diff expressed genes in Males
top50_MLiver_trin_genes <- cbind(rownames(head(liver_con_res_p05[order(
  liver_con_res_p05$log2FoldChange, decreasing = TRUE),], n=50)))
#Run once to create the file and then commented out
#write.table(top50_MLiver_trin_genes,
 #          'SN_maleL_top50TRgenes.txt', 
  #         sep = "", 
   #        quote=FALSE, 
    #         row.names = FALSE, 
     #        col.names = FALSE)

#Pulling the top 50 diff expressed genes in Females
top50_Fliver_trin_genes <- cbind(rownames(head(liver_con_res_p05[order(
  liver_con_res_p05$log2FoldChange, decreasing = FALSE),], n=50)))

#Run once to create the file and then commented out
#write.table(top50_Fliver_trin_genes,
 #           'SN_femL_top50TRgenes.txt', 
  #          sep = "", 
   #         quote=FALSE, 
    #        row.names = FALSE, 
     #       col.names = FALSE)
```

### M-F Gill Comparisson
```{r m-f-gill}
##Pulling out the gill M-F results
gill_con_res <- results(ddsMF_SN_exp, contrast = c("group", "MGill", "FGill"),
                        alpha = 0.05)
gill_con_res$trin_geneid <- rownames(gill_con_res)
head(gill_con_res)

summary(gill_con_res)
```

In the gills there were `r sum(gill_con_res$padj <= 0.05 & gill_con_res$log2FoldChange >= 2, na.rm = TRUE)` genes that we can consider male-biased and `r sum(gill_con_res$padj <= 0.05 & gill_con_res$log2FoldChange <= -2, na.rm = TRUE)` female-biased genes based on our criteria for sex-bias.

I have then pulled out all of the male-biased, female-biased, and non-biased genes to save them into their own objects.

```{r gill-sex-biased}
#Removing the rows where padj. is NA in results
gill_con_res_noNA <- gill_con_res[!is.na(gill_con_res$padj), ]
summary(gill_con_res_noNA) #We can now see that there are no outliers or 
                           #low counts since the NAs have been removed

#Creating a vector that contains all of the male-biased and female-biased genes in the gills
gill_mal_biased <- gill_con_res_noNA[which(gill_con_res_noNA$log2FoldChange >= 2 
                                           & gill_con_res_noNA$padj <= 0.05),]
gill_fem_biased <- gill_con_res_noNA[which(gill_con_res_noNA$log2FoldChange <= -2 
                                           & gill_con_res_noNA$padj <= 0.05),]

#Creating an object that contains all of the non-biased genes in the gills, p>0.05
gill_non_biased <- gill_con_res_noNA[which(gill_con_res_noNA$padj > 0.05),]
```

Here I am getting the trinity gene IDs that correspond to the top 50 differentially expressed genes for males and females. Because there are only 27 female-biased and male-biased genes I will only use those.

```{r top50-gill}
#Creating a subset of results where p-value is less than 0.05
gill_con_res_p05 <- gill_con_res_noNA[gill_con_res_noNA$padj <= 0.05,]

#Pulling the top 27 diff expressed genes in Males
top27_MGill_trin_genes <- cbind(rownames(head(gill_con_res_p05[order(
  gill_con_res_p05$log2FoldChange, decreasing = TRUE),], n=27)))

#Run once to create the file and then commented out
#write.table(top27_MGill_trin_genes,
 #           'SN_maleG_top27TRgenes.txt', 
  #          sep = "", 
   #         quote=FALSE, 
    #        row.names = FALSE, 
     #       col.names = FALSE)

#Pulling the top 27 diff expressed genes in Females 
top27_FGill_trin_genes <- cbind(rownames(head(gill_con_res_p05[order(
  gill_con_res_p05$log2FoldChange, decreasing = FALSE),], n=27)))

#Run once to create the file and then commented out
#write.table(top27_FGill_trin_genes,
 #           'SN_femG_top27TRgenes.txt', 
  #          sep = "", 
   #         quote=FALSE, 
    #        row.names = FALSE, 
     #       col.names = FALSE)
```

### M-F Gonad Comparisson
```{r m-f-gonads}
##Pulling out the gonad M-F results
gonad_con_res <- results(ddsMF_SN_exp, 
                         contrast = c("group", "MGonad", "FGonad"),
                         alpha = 0.05)
gonad_con_res$trin_geneid <- rownames(gonad_con_res)
head(gonad_con_res)

summary(gonad_con_res)
```

Between the ovaries and testes (gonads) there were `r sum(gonad_con_res$padj <= 0.05 & gonad_con_res$log2FoldChange >= 2, na.rm = TRUE)` genes that we can consider male-biased and `r sum(gonad_con_res$padj <= 0.05 & gonad_con_res$log2FoldChange <= -2, na.rm = TRUE)` female-biased genes based on our criteria for sex-bias.

From the gonad M-F results I have then pulled out all of the male-biased, female-biased, and non-biased genes and saved them to their own objects.

```{r gonad-sex-biased}
#Removing the rows where padj. is NA in results
gonad_con_res_noNA <- gonad_con_res[!is.na(gonad_con_res$padj), ]
summary(gonad_con_res_noNA) #We can now see that there are no outliers or 
                            #low counts since the NAs have been removed

#Creating an object that contains all of the male-biased and female-biased genes in the gonads
gonad_mal_biased <- gonad_con_res_noNA[which(gonad_con_res_noNA$log2FoldChange >= 2 
                                             & gonad_con_res_noNA$padj <= 0.05),]
gonad_fem_biased <- gonad_con_res_noNA[which(gonad_con_res_noNA$log2FoldChange <= -2 
                                             & gonad_con_res_noNA$padj <= 0.05),]

#Creating an object that contains all of the non-biased genes in the gonads, p>0.05
gonad_non_biased <- gonad_con_res_noNA[which(gonad_con_res_noNA$padj > 0.05),]
```

I then pulled out the trinity gene IDs that corresponded to the top 50 differentially expressed genes in males and females so that I could BLAST the sequences.
```{r top50-gonad}
#Creating a subset of the results where the p-value is less than 0.05
gonad_con_res_p05 <- gonad_con_res_noNA[gonad_con_res_noNA$padj <= 0.05,]

#Pulling the top 50 diff expressed genes in Males
top50_MGonad_trin_genes <- cbind(rownames(head(gonad_con_res_p05[order(
  gonad_con_res_p05$log2FoldChange, decreasing = TRUE),], n=50)))

#Run once to create the file and then commented out
#write.table(top50_MGonad_trin_genes,
 #           'SN_maleGon_top50TRgenes.txt', 
  #          sep = "", 
   #         quote=FALSE, 
    #        row.names = FALSE, 
     #       col.names = FALSE)

#Pulling the top 50 diff expressed genes in Females
top50_FGonad_trin_genes <- cbind(rownames(head(gonad_con_res_p05[order(
  gonad_con_res_p05$log2FoldChange, decreasing = FALSE),], n=50)))

#Run once to create the file and then commented out
#write.table(top50_FGonad_trin_genes,
 #           'SN_femGon_top50TRgenes.txt', 
  #          sep = "", 
   #         quote=FALSE, 
    #        row.names = FALSE, 
     #       col.names = FALSE)
```

### MA Plots
```{r MA-plot-MF, echo=FALSE, fig.height= 15, fig.width=8, fig.cap="MA-plots generated for each organ type that compares logFC to the mean expression. Female-biased and Male-biased genes are represented by color (green and purple correspondingly) and are determing with a fc cutoff of 2 and a p-value cutoff of 0.05."}
liver_MA <- ggmaplot(liver_con_res,
         fdr = 0.05, fc = 2,
         size = 1.4,
         top = 0,
         main = expression("Male Liver" %->% "Female Liver"),
         font.label = c("bold", 11), label.rectangle = TRUE,
         palette = c("#beaed4", "#7fc97f", "gray27"),
         select.top.method = c("padj", "fc"),
         ggtheme = ggplot2::theme_minimal()
         )

gill_MA <- ggmaplot(gill_con_res,
         fdr = 0.05, fc = 2,
         size = 1.4,
         top = 0,
         main = expression("Male Gill" %->% "Female Gill"),
         font.label = c("bold", 11), label.rectangle = TRUE,
         palette = c("#beaed4", "#7fc97f", "gray27"),
         select.top.method = c("padj", "fc"),
         ggtheme = ggplot2::theme_minimal()
         )

gonad_MA <- ggmaplot(gonad_con_res,
         fdr = 0.05, fc = 2,
         size = 1.4,
         top = 0,
         main = expression("Testis" %->% "Ovaries"),
         font.label = c("bold", 11), label.rectangle = TRUE,
         palette = c("#beaed4", "#7fc97f", "gray27"),
         select.top.method = c("padj", "fc"),
         ggtheme = ggplot2::theme_minimal()
         )

#Merge all the plots into one
plot_grid(liver_MA, gill_MA, gonad_MA,
          ncol = 1,
          nrow = 3,
          labels = c('A', 'B', 'C'),
          label_fontfamily = 'sans',
          label_fontface = 'bold',
          label_size = 18,
          rel_heights = c(4.10, 4.10, 4.10))

```

From the MA plots we can see the highest number of biased genes are found in the gonads (Fig. \@ref(fig:MA-plot-MF)) and then the liver and lastly the gills. The liver appears to be the only organ where there are more female-biased genes than male-biased genes, which is initially surprising due to the sex-role-reversed nature of these fish.

## Creating an Upset Plot
```{r upset-plot-mal, echo=FALSE, fig.cap="Upset plots to show the number of shared sex-biased genes across the organs in males."}
#Male-biased genes
listInputMB <- list("Testis" = rownames(gonad_mal_biased),
                  "Male Gill"=rownames(gill_mal_biased), 
                  "Male Liver"=rownames(liver_mal_biased)
                  )
upset(fromList(listInputMB),
      mainbar.y.label = "Number of Shared Male-Biased Genes",
      sets.x.label = "Total Number of Male-Biased Genes",
      point.size = 3)
```

```{r upset-plot-fem, echo=FALSE, fig.cap="Upset plots to show the number of shared sex-biased genes across the organs in females."}
#Female-biased genes
listInputFB <- list("Ovaries" = rownames(gonad_fem_biased),
                  "Female Gill"=rownames(gill_fem_biased), 
                  "Female Liver"=rownames(liver_fem_biased)
                  )
upset(fromList(listInputFB),
      mainbar.y.label = "Number of Shared Female-Biased Genes",
      sets.x.label = "Total Number of Female-Biased Genes",
      point.size = 3)
```

```{r upset-plot-all, echo=FALSE, fig.cap="Upset plots to show the number of shared sex-biased genes across the organs across both sexes."}
#Exporting data used to make the Upset plot (run once and them commented out)
#write.table(as.data.frame(gonad_mal_biased), "data/gonad_mal_biased.txt", 
      #      row.names = TRUE)
#write.table(as.data.frame(gonad_fem_biased), "data/gonad_fem_biased.txt", 
     #       row.names = TRUE)
#write.table(as.data.frame(gill_mal_biased), "data/gill_mal_biased.txt", 
    #        row.names = TRUE)
#write.table(as.data.frame(gill_fem_biased), "data/gill_fem_biased.txt", 
   #         row.names = TRUE)
#write.table(as.data.frame(liver_mal_biased), "data/liver_mal_biased.txt", 
  #          row.names = TRUE)
#write.table(as.data.frame(liver_fem_biased), "data/liver_fem_biased.txt", 
 #           row.names = TRUE)

#Checking there is no overlap between the two sexes
listInputall <- list("Testis" = rownames(gonad_mal_biased),
                  "Male Gill"=rownames(gill_mal_biased), 
                  "Male Liver"=rownames(liver_mal_biased),
                  "Ovaries" = rownames(gonad_fem_biased),
                  "Female Gill"=rownames(gill_fem_biased), 
                  "Female Liver"=rownames(liver_fem_biased))

upset(fromList(listInputall),
      mainbar.y.label = "Number of Shared Sex-Biased Genes",
      sets.x.label = "Total Number of Sex-Biased Genes",
      point.size = 3,
      nsets = 6,
      nintersects = NA)

```

In both males and females the highest number of shared sex-biased genes is found between the gonads and the liver (Fig. \@ref(fig:upset-plot-mal), \@ref(fig:upset-plot-fem)). There are no genes that are shared across the same organ between males and females (i.e. biased in males and females in the liver) which is good (Fig. \@ref(fig:upset-plot-all)). Two sex-biased genes are female-biased in all organs and three genes are male-biased in all three organs (Fig. \@ref(fig:upset-plot-mal), \@ref(fig:upset-plot-fem)).

## Variation in FC across sex-bias and tissue type
I want to create a plot the highlights the variation we see in fold-change both across the different biases groups (male-biased, female-biased, and non-biased) within one tissue type and across all of the tissue types. To do this I first need to create a "long" dataset in the following format:

| Tissue_type|     Bias      |   Fold-Change   |
|:----------:|:-------------:|:---------------:|
|    Gill    |   male_bias   |      XXXXX      |
|    Gill    |   male_bias   |      XXXXX      |
|    Gill    |   fem_bias    |      XXXXX      |
|    Gill    |   no_bias     |      XXXXX      |
|    ...     |    ....       |       ...       |
|    Liver   |   male_bias   |      XXXXX      |
|    Liver   |   male_bias   |      XXXXX      |
|    Liver   |   fem_bias    |      XXXXX      |
|    Liver   |   no_bias     |      XXXXX      |
|    ...     |    ....       |       ...       |
|   Gonad    |   male_bias   |      XXXXX      |
|   Gonad    |   male_bias   |      XXXXX      |
|   Gonad    |   fem_bias    |      XXXXX      |
|   Gonad    |   no_bias     |      XXXXX      |
|    ...     |    ....       |       ...       |

That dataset is created here, with the Trinity gene IDs included as well:

```{r logFC-long-dataset}
logFC_long<- data.frame(
  tissue=c(rep("Gill",nrow(gill_fem_biased)),
           rep("Gill", nrow(gill_mal_biased)),
           rep("Gill", nrow(gill_non_biased)),
           rep("Gonad", nrow(gonad_fem_biased)),
           rep("Gonad", nrow(gonad_mal_biased)),
           rep("Gonad", nrow(gonad_non_biased)),
           rep("Liver", nrow(liver_fem_biased)),
           rep("Liver", nrow(liver_mal_biased)),
           rep("Liver", nrow(liver_non_biased))
         ),
  bias=c(rep("FB",nrow(gill_fem_biased)),
         rep("MB",nrow(gill_mal_biased)),
         rep("UB", nrow(gill_non_biased)),
         rep("FB", nrow(gonad_fem_biased)),
         rep("MB", nrow(gonad_mal_biased)),
         rep("UB", nrow(gonad_non_biased)),
         rep("FB", nrow(liver_fem_biased)),
         rep("MB", nrow(liver_mal_biased)),
         rep("UB", nrow(liver_non_biased))
         ),
  logFC=c(gill_fem_biased$log2FoldChange,
          gill_mal_biased$log2FoldChange,
          gill_non_biased$log2FoldChange,
          gonad_fem_biased$log2FoldChange,
          gonad_mal_biased$log2FoldChange,
          gonad_non_biased$log2FoldChange,
          liver_fem_biased$log2FoldChange,
          liver_mal_biased$log2FoldChange,
          liver_non_biased$log2FoldChange
          ),
  geneID=c(rownames(gill_fem_biased),
           rownames(gill_mal_biased),
           rownames(gill_non_biased),
           rownames(gonad_fem_biased),
           rownames(gonad_mal_biased),
           rownames(gonad_non_biased),
           rownames(liver_fem_biased),
           rownames(liver_mal_biased),
           rownames(liver_non_biased)
           )
  
)

#Exporting this data, run once and then commented out
#write.csv(logFC_long, "data/logFC_long_sexbias.csv", row.names = FALSE)
```

With the dataset in the proper format now, we can generate the plot
```{r FC-var-plot, echo=FALSE, fig.cap="Absolute value of the logFC for genes that are female-biased, male-biased, and unbiased across each tissue type. Raw fold-change values are added ontop of the boxplot as jitters."}
my_colors <- c("FB" = "#7fc97f", "MB" = "#beaed4", "UB" = "darkgray")

ggplot(logFC_long,
       aes(x = bias, y = abs(logFC), fill = bias)) +
  geom_boxplot() +
  scale_fill_manual(values = my_colors) +
  facet_grid(. ~ tissue) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) +
  geom_point(data=logFC_long, aes(x=bias, y=abs(logFC)),
             size=0.4, 
             position = position_jitter(width = 0.15)) + 
  labs(x="Bias Level", y="|logFC|") +
  guides(fill = guide_legend(title = "Bias Level", order = 3))

```

There appears to be differences in the logFC across both the organs and the sex-bias groups (male-biased, fem-biased, and non-biased) (Fig. \@ref(fig:FC-var-plot)). Let's explore some analyses to see if we can put some statistical power behind those differences.

```{r FC-var-stats, warning=FALSE, eval=FALSE}
#Looking at some summary statistics for logFC between the different groups
tapply(abs(logFC_long$logFC), list(logFC_long$bias, logFC_long$tissue), mean)
tapply(abs(logFC_long$logFC), list(logFC_long$bias, logFC_long$tissue), sd)

par(mfrow=c(1,1))
interaction.plot(logFC_long$tissue, logFC_long$bias, logFC_long$logFC)

logFC_long$tissue <- as.factor(logFC_long$tissue)
logFC_long$bias <- as.factor(logFC_long$bias)

fc_var_aov <- aov(abs(logFC_long$logFC) ~ logFC_long$bias*logFC_long$tissue)
anova(fc_var_aov)

par(mfrow = c(2,2))
plot(fc_var_aov)

library(car)
leveneTest(fc_var_aov)
par(mfrow=c(1,1))
hist(resid(fc_var_aov))
#shapiro.test(resid(fc_var_aov)) 

plot(TukeyHSD(fc_var_aov))
library(multcompView)
multcompLetters4(fc_var_aov, TukeyHSD(fc_var_aov))

##Fitting a negative binomial model
library(MASS)
model <- glm.nb(abs(logFC_long$logFC) ~ logFC_long$tissue * logFC_long$bias)

linear_model <- lm(abs(logFC_long$logFC)~ logFC_long$tissue * logFC_long$bias)
```

## Categorizing sex-specific genes
Now that we have investigated what the sex-bias is like across the different tissue types I want to dive further into genes with **sex-specific expression**. These are genes expressed only in one sex or the other. I am classifying a sex-specific genes within each tissue as ones where the expression of that gene is less than 10 for all of one sex and there is a median of $\ge 20$ in the other sex. 

Prior to running the for loop, any "outliers" that were determined by DESeq were removed from the pool of genes. Additionally, normalized counts were used for the classification of sex-specific genes as DESeq2 uses the moralized counts for all of the logFC/sexbias calculations.

```{r sex-specific-for-loop}
#Pulling out the geneIDs for genes that were categorized as "outliers" by DESeq2
#Calculating the Cooks threshold that would have been used
np <- length(resultsNames(ddsMF_SN_exp))
nsamp <- ncol(ddsMF_SN_exp)
cooks_thresh <- qf(0.99, df1 = np, df2 = nsamp-np)

out_ids <- names(mcols(ddsMF_SN_exp)$maxCooks[mcols(ddsMF_SN_exp)$maxCooks >
                                                cooks_thresh])

#Checking to make sure we have correct number of outliers
length(out_ids)

#Filtering out the dds dataset to remove the outliers determined by DESeq
ddsMF_SN_exp_filtered <- ddsMF_SN_exp[!(rownames(ddsMF_SN_exp) %in% out_ids), ]

#Create a vector with the different organ types
organs <- levels(colData(ddsMF_SN_exp_filtered)$organ)

#Create an empty list to store my datasets in
SN_sex_specific_genes <- list()

#Generate the for loop to identify MSpecific and FSpecific genes in each 
#organ based on Medians
for(organ in organs){
  
  #Male-Specific Genes
  ##Pull out all of the rows where fem count <=10 in every female sample
  fem0_organ_names <- which(rowSums(t(apply(counts(ddsMF_SN_exp_filtered, 
                                                   normalized = TRUE)[, ddsMF_SN_exp_filtered$sex == "F" 
                                                                      & 
                                                                        ddsMF_SN_exp_filtered$organ == organ], 
                                            1, 
                                            function(x) x <= 10)
                                      )
                                ) == ncol(counts(ddsMF_SN_exp_filtered, 
                                                 normalized = TRUE)[, ddsMF_SN_exp_filtered$sex == "F" &
                                                                      ddsMF_SN_exp_filtered$organ == organ])
                      )
  
  fem0_organ <- counts(ddsMF_SN_exp_filtered, 
                       normalized = TRUE)[rownames(counts(ddsMF_SN_exp_filtered, 
                                                          normalized = TRUE)) %in% 
                                            names(fem0_organ_names), 
                                          ddsMF_SN_exp_filtered$organ == organ]
                                                         
  
  ##Pull out rows where median of male count >=20 for that organ
  mal10_organ <- apply(counts(ddsMF_SN_exp_filtered, 
                              normalized = TRUE)[, ddsMF_SN_exp_filtered$sex == "M" 
                                                 & ddsMF_SN_exp_filtered$organ == organ], 
                       1,
                       function(x) median(x) >=20)
  
  ##Keep only the rows where all fem samples <=10 and the Male median>=20
  fem0_mal10_organ <- fem0_organ[rownames(fem0_organ) %in% 
                                   names(mal10_organ[mal10_organ == TRUE]),
                                 ]
  
  ##Create a new object with a name based on the organ type
  organ_malsp <- sub("$", "_male_specific", organ)
  SN_sex_specific_genes[[organ_malsp]] <- fem0_mal10_organ
  
  
  
  #Female-Specific Genes
  ##Pull out all of the rows where male count <=10 in every male sample
  mal0_organ_names <- which(rowSums(t(apply(counts(ddsMF_SN_exp_filtered, 
                                                   normalized = TRUE)[, ddsMF_SN_exp_filtered$sex == "M" 
                                                                      & 
                                                                        ddsMF_SN_exp_filtered$organ == organ],
                                            1, 
                                            function(x) x <= 10)
                                      )
                                    ) == ncol(counts(ddsMF_SN_exp_filtered, 
                                                     normalized = TRUE)[, ddsMF_SN_exp_filtered$sex == "M" & 
                                                                          ddsMF_SN_exp_filtered$organ == organ])
                            )
  
  mal0_organ <- counts(ddsMF_SN_exp_filtered, 
                       normalized = TRUE)[rownames(counts(ddsMF_SN_exp_filtered, 
                                                          normalized = TRUE)) %in% 
                                            names(mal0_organ_names), 
                                          ddsMF_SN_exp_filtered$organ == organ]
  
  ##Pull out rows where median of female count >=10 for that organ
  fem10_organ <- apply(counts(ddsMF_SN_exp_filtered, 
                              normalized = TRUE)[, ddsMF_SN_exp_filtered$sex == "F" &
                                                   ddsMF_SN_exp_filtered$organ == organ],
                       1,
                       function(x) median(x) >=20)
  
  #Keep only the rows where male=0 and the fem median>=10
  mal0_fem10_organ <- mal0_organ[rownames(mal0_organ) %in% 
                                   names(fem10_organ[fem10_organ == TRUE]),
                                 ]
  
  # Create a new object with a name based on the organ type
  organ_femsp <- sub("$", "_female_specific", organ)
  SN_sex_specific_genes[[organ_femsp]] <- mal0_fem10_organ
  
} 
```

### Investigating the results of the sex-specific subsetting
Let's now take a look at how many sex-specific genes we have in each tissue type:
```{r sex-specific-counts, echo=FALSE, message=FALSE, warning=FALSE}
sex_specific_counts <- data.frame(tissue = rep(c("Gill", "Gonad", "Liver"), 2),
                                  sex = c(rep("M", 3),
                                          rep("F", 3)),
                                  num_genes = c(nrow(SN_sex_specific_genes$Gill_male_specific),
                                                nrow(SN_sex_specific_genes$Gonad_male_specific),
                                                nrow(SN_sex_specific_genes$Liver_male_specific),
                                                nrow(SN_sex_specific_genes$Gill_female_specific),
                                                nrow(SN_sex_specific_genes$Gonad_female_specific),
                                                nrow(SN_sex_specific_genes$Liver_female_specific))
                                  )

sex_specific_counts %>% 
  group_by(tissue, sex) %>% 
  summarise("Number of Sex-specific Genes" = sum(num_genes)) %>%
  kable(caption = "Sex-specific Genes", format = "html",
        col.names = c("Tissue Type", "Sex", "Number of Sex-specific Genes")) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left")
```

To get a better idea what is going on between the sex-specific genes vs the sex-biased genes I first determined how many overlaps there were between our sex-biased and sex-specific genes and then used plotCounts to plot the counts for each gene separately. 

In the **Male Gills** there are `r nrow(SN_sex_specific_genes$Gill_male_specific[rownames(SN_sex_specific_genes$Gill_male_specific) %in% rownames(gill_mal_biased),])` genes that overlap between sex-biased and sex-specific, there are `r nrow(SN_sex_specific_genes$Gonad_male_specific[rownames(SN_sex_specific_genes$Gonad_male_specific) %in% rownames(gonad_mal_biased),])` overlapping in the **gonads** and `r nrow(SN_sex_specific_genes$Liver_male_specific[rownames(SN_sex_specific_genes$Liver_male_specific) %in% rownames(liver_mal_biased),])` overlapping in the **liver**. For females we have `r nrow(SN_sex_specific_genes$Gill_female_specific[rownames(SN_sex_specific_genes$Gill_female_specific) %in% rownames(gill_fem_biased),])` overlapping sex-specific and sex-biased genes in the **gills**, `r nrow(SN_sex_specific_genes$Gonad_female_specific[rownames(SN_sex_specific_genes$Gonad_female_specific) %in% rownames(gonad_fem_biased),])` genes in the **gonads**, and `r nrow(SN_sex_specific_genes$Liver_female_specific[rownames(SN_sex_specific_genes$Liver_female_specific) %in% rownames(liver_fem_biased),])` genes in the **liver**.

```{r gills-sex-specific, eval = FALSE}
## Female sex-specific
fem_sp_geneIDs <- rownames(SN_sex_specific_genes$Gill_female_specific)

##Set the plotting window
par(mfrow=c(1,1))

for(gene in fem_sp_geneIDs){
  
  plotCounts(ddsMF_SN[,ddsMF_SN$organ == 'Gill'], gene=gene, intgroup="sex")
  
}

## Female sex-biased
gill_fem_biased_geneIDS <- rownames(gill_fem_biased)

##Set the plotting window
par(mfrow=c(1,1))

for(gene in gill_fem_biased_geneIDS){
  
  plotCounts(ddsMF_SN[,ddsMF_SN$organ == 'Gill'], gene=gene, intgroup="sex", col= 'salmon', pch = 19)
  
}
```

### Pulling out the gene IDs for all of the sex-specific genes
```{r}
#Run once to create the file and then commented out
#write.table(cbind(rownames(SN_sex_specific_genes$Gill_male_specific)),
 #           'SN_malG_specific_TRgenes.txt', 
  #          sep = "", 
   #         quote=FALSE, 
    #        row.names = FALSE, 
     #       col.names = FALSE)
#write.table(cbind(rownames(SN_sex_specific_genes$Gill_female_specific)),
 #           'SN_femG_specific_TRgenes.txt', 
  #          sep = "", 
   #         quote=FALSE, 
    #        row.names = FALSE, 
     #       col.names = FALSE)
#write.table(cbind(rownames(SN_sex_specific_genes$Gonad_male_specific)),
 #           'SN_malGon_specific_TRgenes.txt', 
  #          sep = "", 
   #         quote=FALSE, 
    #        row.names = FALSE, 
     #       col.names = FALSE)
#write.table(cbind(rownames(SN_sex_specific_genes$Gonad_female_specific)),
 #           'SN_femGon_specific_TRgenes.txt', 
  #          sep = "", 
   #         quote=FALSE, 
    #        row.names = FALSE, 
     #       col.names = FALSE)
#write.table(cbind(rownames(SN_sex_specific_genes$Liver_male_specific)),
 #           'SN_malL_specific_TRgenes.txt', 
  #          sep = "", 
   #         quote=FALSE, 
    #        row.names = FALSE, 
     #       col.names = FALSE)
#write.table(cbind(rownames(SN_sex_specific_genes$Liver_female_specific)),
 #           'SN_femL_specific_TRgenes.txt', 
  #          sep = "", 
   #         quote=FALSE, 
    #        row.names = FALSE, 
     #       col.names = FALSE)
```

## Creating categories and binning the sex-biased genes based on degree of logFC
This binned was only done on the sex-biased genes, will not have a category for the unbiased genes. The cutoffs for the different groups are as follows:

    1. Low - biased = LFC 2 - 3
    2. Medium - biased = LFC 3 - 5
    3. High - biased = LFC 5 - 10
    4. Extreme sex bias = LFC > 10

Counts of the sex-specific genes will be added on separately as they were not classified based on fold-change but rather a presence in one sex and an absence in the other sex.

```{r bin-sex-bias}
#Make a vector that contains all of the groupings
biased_bins <- c("Unbiased", "Low", "Med", "High", "Extreme", "Sex-specific")

#Create a new column in the dataset and use ifelse statements to set the category limits
#abs(logFC) was used to account for the fem-biased genes possessing negative values
logFC_long$bias_cat <- ifelse(logFC_long$bias == "UB",
                              biased_bins[1],
                              ifelse(abs(logFC_long$logFC) >= 2 & abs(logFC_long$logFC) < 3,
                                     biased_bins[2],
                                     ifelse(abs(logFC_long$logFC) >= 3 & abs(logFC_long$logFC) < 5,
                                            biased_bins[3],
                                            ifelse(abs(logFC_long$logFC) >= 5 & abs(logFC_long$logFC) < 10,
                                                   biased_bins[4],
                                                   biased_bins[5])
                                            )
                                     )
                              )

#Making sure the genes we categorized as sex-specific are labeled as sex-specific for their 
#bias cat. in the dataset
organs <- c("Gill", "Gill", "Gonad", "Gonad", "Liver", "Liver")
bias <- c("MB", "FB", "MB", "FB", "MB", "FB")

for(i in 1:length(SN_sex_specific_genes)){

  tmp <- SN_sex_specific_genes[[i]]
  tmp <- as.data.frame(tmp)
  tmp$geneID <- rownames(tmp)
  
  for(j in 1:nrow(tmp)){
    
    one_gene <- tmp[j, ]
    
    if(one_gene[["geneID"]] %in%
       logFC_long[logFC_long$tissue == organs[[i]] &
                  logFC_long$bias == bias[[i]],"geneID"]){
       
      logFC_long[logFC_long$geneID == one_gene[["geneID"]] &
                   logFC_long$tissue == organs[[i]] &
                   logFC_long$bias == bias[[i]],
                 "bias_cat"] <- "Sex-specific"
    }else{
      
      one_gene_dat <- data.frame(matrix(ncol= ncol(logFC_long),
                                        nrow=1))
      colnames(one_gene_dat) <- colnames(logFC_long)
      
      one_gene_dat$tissue <- organs[[i]]
      one_gene_dat$geneID <- one_gene[["geneID"]]
      one_gene_dat$bias <- bias[[i]]
      one_gene_dat$bias_cat <- "Sex-specific"
      rownames(one_gene_dat) <- NULL
      
      logFC_long <- rbind(one_gene_dat, logFC_long)
      
      rownames(logFC_long) <- NULL
    }
  }
 }

#Create  subset of our long dataset that does not include the non-biased genes
logFC_long_noNB <- logFC_long[logFC_long$bias_cat != "Unbiased",]

#Make a table to count the number of genes in each category for each organ
bias_cat_table <- table(logFC_long_noNB$bias, logFC_long_noNB$bias_cat, logFC_long_noNB$tissue)

#Add the counts of sex-specific genes to the table
bias_cat_gill <- bias_cat_table[,, "Gill"]
#write.table(bias_cat_gill, "data/bias_cat_gills.txt", row.names = TRUE)

bias_cat_gonad <- bias_cat_table[,, "Gonad"]
#write.table(bias_cat_gonad, "data/bias_cat_gonad.txt", row.names = TRUE)

bias_cat_liver <- bias_cat_table[,, "Liver"]
write.table(bias_cat_liver, "data/bias_cat_liver.txt", row.names = TRUE)

#Plot the counts for each tissue type
par(mfrow=c(1, 3))
barplot(bias_cat_gill[,biased_bins[biased_bins!="Unbiased"]], beside = TRUE, ylim = c(0, max(bias_cat_table) + 1), col = c("#7fc97f", "#beaed4"),
        main = "Gill", ylab = "Number of Genes", xlab = "Bias Category")
barplot(bias_cat_gonad[,biased_bins[biased_bins!="Unbiased"]], beside = TRUE, ylim = c(0, max(bias_cat_table) + 1), col = c("#7fc97f", "#beaed4"),
        main = "Gonad")
barplot(bias_cat_liver[,biased_bins[biased_bins!="Unbiased"]], beside = TRUE, ylim = c(0, max(bias_cat_table) + 1), col = c("#7fc97f", "#beaed4"),
        main = "Liver")
legend("topright", legend = c("Female-biased", "Male-biased"), fill = c("#7fc97f", "#beaed4"))

```

## Gene Ontology Analysis
We wanted to figure out two things:   

  1) What are the genes that are sex-biased within each tissue type   
  2) What are the GO terms for those genes.   

To accomplish this I first had to pull out the sequences that corresponded to the Trinity gene IDS pulled out above. Once those sequences were fetched they were blasted against the _Corythoichthys intestinalis_ genome. This was completed with the following bash script on the RCC. 

### Using BLAST in command line
```{bash, eval = FALSE}
#Create a BLAST database - Used Corythoichthys intestinalis for our BLAST database
makeblastdb -in /home/rccuser/shared/emily_files/S_nigra_2024/c_intestinalis_data/data/GCF_030265065.1/GCF_030265065.1_ASM3026506v1_genomic.fna -out c_intestinalis_genome -dbtype nucl
```


```{bash, eval = FALSE}
#!/bin/bash

#Create the arguements
input_dir_TR=$1 #Location of the .txt files that contain the trinity gene IDs
subset_fasta=$2 #Path to the subset_fasta_file script
assembly_file=$3 #Name/location of the de novo assembly
blast_database=$4
output_dir=$5 #Desired output directory for .fasta and BLAST files

## Loop through all the Trinity gene ID .txt files
for file in $input_dir_TR/*TRgenes.txt
        do

        #Extract the sample name from the file name
        sample=$(basename $file .txt)

        #Get the corresponding sequences for the Trinity Gene IDs
        echo "Running subset_fasta_file for ${sample}..."
        $subset_fasta -c -f $assembly_file -l $input_dir_TR/${sample}.txt

        #Rename the automated output from the subset_fasta_file script
        fasta_out=$(basename $assembly_file fasta)
        mv $fasta_out.subset.fasta ${sample}.fasta

        #Blast the sequences
        echo "Running BLAST for ${sample}..."
        blastn -db $blast_database -query ${sample}.fasta -out ${sample}_corythoicthys_blast.txt \
                -evalue 0.001 \
                -num_threads 12 \
                -outfmt "6 qseqid qstart qend stitle salltitles sseqid sallseqid sacc sblastnames sstart send evalue bitscore length pident gaps"       

        #Move the outputs into desired output directory
        mv ${sample}* $output_dir

done

```

This script was run as `bash ../scripts/blast_pipeline.sh nigra_high_exp_genes/TRgenes/ ../scripts/subset_fasta_file nigra_supertranscript.fasta c_intestinalis_genome/c_intestinalis_genome nigra_high_exp_genes/high_exp_results`

Once the BLAST results were obtained in the RCC I read them into R for further filtering (keeping only one hit per gene)
```{r readin_blast_results}
#Specify the directory where BLAST results are located
blast_path <- "data/nigra_BLAST/"

#Create a list of the files I want
SN_blast_files <- list.files(blast_path, pattern = "cory")

#Create an empty list to store my datasets in
SN_blast_list <- list()

#Create a loop to read in all of the blast results
for(file_name in SN_blast_files){
   # Read the file and store it in a new object
  file_data <- read.delim(file.path("data/nigra_BLAST", file_name), 
                          stringsAsFactors = FALSE, 
                          quote = "", 
                          sep = "\t",
                          header = FALSE)
 
  #Add column names to the dataset
  colnames(file_data) <- c("trin_geneid", 
                           "trin_gene_start", 
                           "trin_gene_end", 
                           "cory_gene_info", 
                           "cory_gene_start
                           cory_gene_end", 
                           "evalue", 
                           "bit_score", 
                           "length", 
                           "pident", 
                           "gaps")
  
  # Create a new object with a name based on the file name
  blast_name <- sub(".txt$", "", file_name) #Removes the file extension
  SN_blast_list[[blast_name]] <- file_data
}

```

### Filtering Blast results
Because of the way BLAST was run in the RCC, we can have multiple hits for one gene, I want to filter that out so that there is only one hit kept for each gene.

```{r filter-BLAST}
#Use lapply to apply the function to each dataset stored in the list created above
blast_output_filtered <- lapply(SN_blast_list, function(data_frame){
 
  #Pull out the Unique Trinity gene IDs
  uniqueID <- unique(data_frame[1])
  
  #Create an empty dataframe to store intermediate results in
  output <- data.frame(matrix(data = NA, ncol = ncol(data_frame), nrow = 0))
  colnames(output) <- c("trin_geneid", "trin_gene_start", "trin_gene_end", "cory_gene_info", "cory_gene_start", "cory_gene_end", "evalue", "bit_score", "length", "pident", "gaps")
  
  #Generate a for loop that pulls out the lowest e-value + highest % identity for each gene
  for(gene in uniqueID$trin_geneid){
    
    #Subset the dataset for each Trinity gene
    this_trin_gene<- subset(data_frame, trin_geneid==gene)
    #Pull out gene with smallest e-value
    uni_gene_subset_lowe <- this_trin_gene[which(this_trin_gene$evalue == min(this_trin_gene$evalue)),]
    #In case mult. genes have same e-value, pull out gene with highest % identity
    uni_gene_subset_lowe_highpid <- uni_gene_subset_lowe[which(uni_gene_subset_lowe$pident == max(uni_gene_subset_lowe$pident)),]
    
    # keep only one of multiple identical rows
    uni_gene_subset_lowe_highpid <- unique(uni_gene_subset_lowe_highpid)

    # check that there is a single gene ID in cory, if not, only first row is kept
    if(length(gsub("^.*(GeneID:\\d+)\\].*$","\\1",uni_gene_subset_lowe_highpid$cory_gene_info))>1){
    #   browser()
    # } else{
      uni_gene_subset_lowe_highpid<-uni_gene_subset_lowe_highpid[1,]
    }
    
    #Add the final gene into the empty dataframe from above
    output <- rbind(output, uni_gene_subset_lowe_highpid)
  }
  
  return(output)
})

```

### Merging the BLAST filtered Data frames with the sex-biased information
Not all of the trinity genes had a BLAST hit come back, but I still want to include what those genes are in the dataset. To do this I am merging all of the top 50 genes with the filtered BLAST dataset and if there are genes that did not have any hits (i.e. are not in the filtered BLAST database) then they will receive an "NA".

All genes that had an "NA" come back from blasting against the _C. intestinalis_ database will be blasted against the non-redundant BLAST database as well.

```{r merge-blast-results}
#Getting the top 50 diff. expressed genes in Males
top50_MGonad <- head(gonad_con_res_p05[order(gonad_con_res_p05$log2FoldChange, decreasing = TRUE),], 
                     n=50)
top50_MGill <- head(gill_con_res_p05[order(gill_con_res_p05$log2FoldChange, decreasing = TRUE),], 
                     n=50)
top50_MLiver <- head(liver_con_res_p05[order(liver_con_res_p05$log2FoldChange, decreasing = TRUE),], 
                     n=50)

#Pulling the top 50 diff expressed genes in Females
top50_FGonad <- head(gonad_con_res_p05[order(gonad_con_res_p05$log2FoldChange, decreasing = FALSE),],
                     n=50)
top21_FGill <- head(gill_con_res_p05[order(gill_con_res_p05$log2FoldChange, decreasing = FALSE),],
                     n=21)
top50_FLiver <- head(liver_con_res_p05[order(liver_con_res_p05$log2FoldChange, decreasing = FALSE),],
                     n=50)
#Create dataframes for all of the above that contains the trinity gene ID and the logFoldChange
top50_MGonad_df <- as.data.frame(cbind(trin_geneid = top50_MGonad$trin_geneid,
                                       LogFC = top50_MGonad$log2FoldChange))
top50_MGill_df <- as.data.frame(cbind(trin_geneid = top50_MGill$trin_geneid,
                                       LogFC = top50_MGill$log2FoldChange))
top50_MLiver_df <- as.data.frame(cbind(trin_geneid = top50_MLiver$trin_geneid,
                                       LogFC = top50_MLiver$log2FoldChange))
top50_FGonad_df <- as.data.frame(cbind(trin_geneid = top50_FGonad$trin_geneid, 
                                      LogFC = top50_FGonad$log2FoldChange))
top21_FGill_df <- as.data.frame(cbind(trin_geneid = top21_FGill$trin_geneid, 
                                      LogFC = top21_FGill$log2FoldChange))
top50_FLiver_df <- as.data.frame(cbind(trin_geneid = top50_FLiver$trin_geneid, 
                                      LogFC = top50_FLiver$log2FoldChange))

#Merge the above dataframes with the corresponding blast_output_filtered dataframe
blast_output_filtered$SN_femG_top21TRgenes_blast <- merge(top21_FGill_df, blast_output_filtered$SN_femG_top27TRgenes_corythoicthys_blast, by = "trin_geneid", all.x = TRUE, all.y = TRUE)
blast_output_filtered$SN_femGon_top50TRgenes_blast <- merge(top50_FGonad_df, blast_output_filtered$SN_femL_top50TRgenes_corythoicthys_blast, by = "trin_geneid", all.x = TRUE, all.y = TRUE)
blast_output_filtered$SN_femL_top50TRgenes_blast <- merge(top50_FLiver_df, blast_output_filtered$SN_femL_top50TRgenes_corythoicthys_blast, by = "trin_geneid", all.x = TRUE, all.y = TRUE)
#I need to make top 50 maleG, but while I'm figuring out what went wrong, I'll plug in top 27
blast_output_filtered$SN_maleG_top27TRgenes_blast <- merge(top50_MGill_df, blast_output_filtered$SN_maleG_top27TRgenes_corythoicthys_blast, by = "trin_geneid", all.x = TRUE, all.y = TRUE)
blast_output_filtered$SN_maleGon_top50TRgenes_blast <- merge(top50_MGonad_df, blast_output_filtered$SN_maleGon_top50TRgenes_corythoicthys_blast, by = "trin_geneid", all.x = TRUE, all.y = TRUE)
blast_output_filtered$SN_maleL_top50TRgenes_blast <- merge(top50_MLiver_df, blast_output_filtered$SN_maleL_top50TRgenes_corythoicthys_blast, by = "trin_geneid", all.x = TRUE, all.y = TRUE)
```

### Blasting the sex-specific genes against the _C. intestinalis_ genome
The trinity genes that were pulled out for the sex-specific genes above were blasted with `blastn` against the _C. intestinalis_ genome in the RCC. I then need to load those BLAST results into R, filter them, and pull out what the _C. intestinalis_ gene names are that correspond to them. 

First, I read in all of the BLAST results for the sex-specific genes in each organ.

```{r readin_blast_results_SS}
#Specify the directory where BLAST results are located
blast_path <- "data/nigra_BLAST/"

#Create a list of the files I want
SN_blast_files <- list.files(blast_path, pattern = "cory")

#Create an empty list to store my datasets in
SN_blast_list_sex_specific <- list()

#Create a loop to read in all of the blast results
for(file_name in SN_blast_files){
  # Read the file and store it in a new object
  file_data <- read.delim(file.path(blast_path, file_name), 
                          stringsAsFactors = FALSE, 
                          quote = "", 
                          sep = "\t",
                          header = FALSE)
  
  file_data <- file_data[, c(1:4, 10:16)]
  
  #Add column names to the dataset
  colnames(file_data) <- c("trin_geneid", 
                           "trin_gene_start", "trin_gene_end", 
                           "cory_gene_info",
                           "cory_gene_start", "cory_gene_end", 
                           "evalue", "bit_score", "length", "pident", 
                           "gaps")

  # Create a new object with a name based on the file name
  blast_name <- sub(".txt$", "", file_name) #Removes the file extension
  SN_blast_list_sex_specific[[blast_name]] <- file_data
}

```

I then need to filter to make sure each trinity gene only has one hit in the same way I did above.

```{r filter-BLAST-SS}
#Use lapply to apply the function to each dataset stored in the list created above
blast_output_filtered_SS <- lapply(SN_blast_list_sex_specific, function(data_frame){
 
  #Pull out the Unique Trinity gene IDs
  uniqueID <- unique(data_frame[1])
  
  #Create an empty dataframe to store intermediate results in
  output <- data.frame(matrix(data = NA, ncol = ncol(data_frame), nrow = 0))
  colnames(output) <- c("trin_geneid", "trin_gene_start", "trin_gene_end",
                        "cory_gene_info", "cory_gene_start", "cory_gene_end", 
                        "evalue", "bit_score", "length", "pident", "gaps")
  
  #Generate a for loop that pulls out the lowest e-value + highest % identity for each gene
  for(gene in uniqueID$trin_geneid){
    
    #Subset the dataset for each Trinity gene
    this_trin_gene<- subset(data_frame, trin_geneid==gene)
    #Pull out gene with smallest e-value
    uni_gene_subset_lowe <- this_trin_gene[which(this_trin_gene$evalue ==
                                                   min(this_trin_gene$evalue)),]
    #In case mult. genes have same e-value, pull out gene with highest % identity
    uni_gene_subset_lowe_highpid <- 
      uni_gene_subset_lowe[which(uni_gene_subset_lowe$pident ==
                                   max(uni_gene_subset_lowe$pident)),]
    
    # keep only one of multiple identical rows
    uni_gene_subset_lowe_highpid <- unique(uni_gene_subset_lowe_highpid)

    # check that there is a single gene ID in scovelli, if not, only first row is kept
    if(length(gsub("^.*(GeneID:\\d+)\\].*$",
                   "\\1",
                   uni_gene_subset_lowe_highpid$scov_gene_info))>1){
      
      uni_gene_subset_lowe_highpid <- uni_gene_subset_lowe_highpid[1,]
    }
    
    #Add the final gene into the empty dataframe from above
    output <- rbind(output, uni_gene_subset_lowe_highpid)
  }
  
  return(output)
})

```

Now I am going to pull out the gene name and protein ID from the "cory_gene_info" column.

```{r}
for (i in 1:length(blast_output_filtered_SS)) {
  
  #Pulling out the gene name and adding it to a new column
  blast_output_filtered_SS[[i]]$gene <- gsub("^(.*)(gene=)(\\w+\\d*)(.*$)",
                                             "\\3",
                                             blast_output_filtered_SS[[i]]$cory_gene_info)
  
  #Pulling out the protein name and adding it to a new column
  blast_output_filtered_SS[[i]]$protein <- gsub("^(.*)(protein=)(.*)([[:graph:]]\\s[[:graph:]])(protein_id=)(.*$)",
                                                "\\3",
                                                blast_output_filtered_SS[[i]]$cory_gene_info)
  
  
}

```

```{r long_BLAST}
#Create a long format dataset that has the tissue information, BLAST gene and protein, 
#geneID, and sex 
blast_long_SS <- data.frame(
  tissue=c(rep("Gill", 
               sum(nrow(blast_output_filtered_SS$SN_femG_specific_TRgenes_corythoicthys_blast),
                   nrow(blast_output_filtered_SS$SN_malG_specific_TRgenes_corythoicthys_blast))),
           rep("Gonad", 
               sum(nrow(blast_output_filtered_SS$SN_femGon_specific_TRgenes_corythoicthys_blast),
                   nrow(blast_output_filtered_SS$SN_malGon_specific_TRgenes_corythoicthys_blast))),
           rep("Liver",
               sum(nrow(blast_output_filtered_SS$SN_femL_specific_TRgenes_corythoicthys_blast),
                   nrow(blast_output_filtered_SS$SN_malL_specific_TRgenes_corythoicthys_blast)))
         ),
  sex=c(rep("Female",
            nrow(blast_output_filtered_SS$SN_femG_specific_TRgenes_corythoicthys_blast)),
        rep("Male",
            nrow(blast_output_filtered_SS$SN_malG_specific_TRgenes_corythoicthys_blast)),
        rep("Female",
            nrow(blast_output_filtered_SS$SN_femGon_specific_TRgenes_corythoicthys_blast)),
        rep("Male",
            nrow(blast_output_filtered_SS$SN_malGon_specific_TRgenes_corythoicthys_blast)),
        rep("Female",
            nrow(blast_output_filtered_SS$SN_femL_specific_TRgenes_corythoicthys_blast)),
        rep("Male",
            nrow(blast_output_filtered_SS$SN_malL_specific_TRgenes_corythoicthys_blast))
        ),
  gene=c(blast_output_filtered_SS$SN_femG_specific_TRgenes_corythoicthys_blast$gene,
         blast_output_filtered_SS$SN_malG_specific_TRgenes_corythoicthys_blast$gene,
         blast_output_filtered_SS$SN_femGon_specific_TRgenes_corythoicthys_blast$gene,
         blast_output_filtered_SS$SN_malGon_specific_TRgenes_corythoicthys_blast$gene,
         blast_output_filtered_SS$SN_femL_specific_TRgenes_corythoicthys_blast$gene,
         blast_output_filtered_SS$SN_malL_specific_TRgenes_corythoicthys_blast$gene
         ),
  protein=c(blast_output_filtered_SS$SN_femG_specific_TRgenes_corythoicthys_blast$protein,
            blast_output_filtered_SS$SN_malG_specific_TRgenes_corythoicthys_blast$protein,
            blast_output_filtered_SS$SN_femGon_specific_TRgenes_corythoicthys_blast$protein,
            blast_output_filtered_SS$SN_malGon_specific_TRgenes_corythoicthys_blast$protein,
            blast_output_filtered_SS$SN_femL_specific_TRgenes_corythoicthys_blast$protein,
            blast_output_filtered_SS$SN_malL_specific_TRgenes_corythoicthys_blast$protein
            ),
  trinityID=c(blast_output_filtered_SS$SN_femG_specific_TRgenes_corythoicthys_blast$trin_geneid,
              blast_output_filtered_SS$SN_malG_specific_TRgenes_corythoicthys_blast$trin_geneid,
              blast_output_filtered_SS$SN_femGon_specific_TRgenes_corythoicthys_blast$trin_geneid,
              blast_output_filtered_SS$SN_malGon_specific_TRgenes_corythoicthys_blast$trin_geneid,
              blast_output_filtered_SS$SN_femL_specific_TRgenes_corythoicthys_blast$trin_geneid,
              blast_output_filtered_SS$SN_malL_specific_TRgenes_corythoicthys_blast$trin_geneid)
  )

write.csv(blast_long_SS, "SN_BLAST_results_sex_specific.csv", row.names = FALSE)
```



