---
title: "Analysing *Stigmatopora macropterygia* RNAseq Data"
author: "Emily Beasley"
date: "`r Sys.Date()`"
output: 
    html_document:
        code_folding: show
        toc: yes
        toc_float: yes
        number_sections: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir='../',fig_path="../imgs/")
```

``` {r library, include = FALSE}
#library(tximport) #This was used to generate an abundance matrix from the salmon quant outputs

```
This pipeline follows a pipeline outlined for _Syngnathus floridae_. Changes in scripts and outputs are outlined in this document. This pipeline was run on the University of Canterbury's RCC.

# Pre-assembly Quality Control and Filtering

## Trimming the Reads with Trimmomatic
[Trimmomatic](http://www.usadellab.org/cms/index.php?page=trimmomatic) is commonly used for Illumina paired-end and single ended data. There are a lot of different trimming steps that can be performed and parameters that correspond to each step. 

### Running trimmomatic across Multiple Pairs of Reads

```{bash}
#!/bin/bash

data=$1

for fq in ${data}*_R1_001.fastq.gz
        do
        base=$(basename $fq _R1_001.fastq.gz)
        echo "Running trimmomatic for ${base}..."
        time trimmomatic PE -threads 16 $fq ${data}${base}_R2_001.fastq.gz \
                ${data}trimmed/${base}_paired_R1_001.fastq.gz ${data}trimmed/${base}_unpaired_R1_001.fastq.gz \
                ${data}trimmed/${base}_paired_R2_001.fastq.gz ${data}trimmed/${base}_unpaired_R2_001.fastq.gz \
                ILLUMINACLIP:TruSeq3-PE-2.fa:2:30:10 HEADCROP:12 LEADING:3 TRAILING:10 SLIDINGWINDOW:4:15 MINLEN:50
done

```



-   Before running the script, change the `trimmomatic` line to `echo "..."` to make sure all of the variables are working correctly.

-   Trimmomatic takes about a day to run, so either run it with nohup or screen 

-   Then remove the `echo "..."` and run the script as `nohup bash ../../scripts/trimmomatic_run.sh ../ > trim.out 2>&1 &`.

## Using SortMeRNA to remove rRNA contamination

### Creating a for loop to run multiple samples at a time
```{bash}
#!/bin/bash

#Create arguments
input_dir=$1 #location of the reads
ref_fasta=$2 #desired reference fasta
output_dir_rrna=$3 #desired location for the reads that are rRNA
output_dir_norrna=$4 #desired location for the reads that are NOT rRNA

## Loop through all pairs of reads in the input directory
for pair in $1/*_R1_001.fastq.gz
        do

        #Extract the sample name from the file name
        sample=$(basename $pair _R1_001.fastq.gz)

        #Extract Fish ID
        ID=$(basename $pair _paired_R1_001.fastq.gz)

        ##Echo the sample name it is currently running
        echo "Running SortMeRNA for ${ID}..."

        #Define the paths to the input and output files for this sample
        input1=$1/${sample}_R1_001.fastq.gz
        input2=$1/${sample}_R2_001.fastq.gz

        #Run SortMeRNA on this pair of reads
        time sortmerna --threads 16 --ref $2 --reads $input1 --reads $input2 --fastx --aligned $3/${ID} --other $4/${ID} --out2

        #Remove SortMeRNA intermediate files before running again
        rm -r /home/rccuser/sortmerna/run/kvdb

done


```

This script was run as `nohup bash ../../scripts/sortmerna_run.sh paired/ /home/rccuser/shared/rRNA_databases_v4.3.4/smr_v4.3_fast_db.fasta rrna/ no_rrna/ > sortmerna.log 2>&1 &`