---
title: "WGCNA and cluster evolution in _S. macropterygia_"
author: Sarah P. Flanagan
output:
  bookdown::pdf_document2:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    toc: true
header-includes: >
  \usepackage{lipsum}
  \usepackage{float}
  \floatplacement{figure}{H}
editor_options:
  chunk_output_type: console
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```
```{r setdir}
knitr::opts_knit$set(root.dir='../')
knitr::opts_chunk$set(fig.path = "../figs/",
                      dev="png",
                      dpi=300)
```


``` {r library, include = TRUE, message = FALSE, warning = FALSE}
#This is a comprehensive list of all the libraries used in this document
library(knitr)
library(kableExtra)
library(spfTools)
library(WGCNA)
library(vioplot)
library(multcompView)
library(RColorBrewer)
sex_cols <- c("F" = "#7fc97f", "M" = "#beaed4")
```

``` {r traitData, eval=FALSE}
#The samples file generated for tximport
samples <- read.table("data/SM_samples.txt", header = TRUE)

#Make sure the conditions are in the samples file as a factor
samples$sex <- as.factor(samples$sex)
samples$organ <- as.factor(samples$organ)
trait_data<-data.frame(
  female=ifelse(samples$sex=="F",1,0),
  gill=ifelse(samples$organ=="Gill",1,0),
  gonad=ifelse(samples$organ=="Gonad",1,0),
  liver=ifelse(samples$organ=="Liver",1,0)
)
rownames(trait_data)<-samples$ID
write.csv(trait_data,"data/trait_data_for_WGCNA.csv",row.names = TRUE)
```
```{r readData}
trait_data<-read.csv("data/trait_data_for_WGCNA.csv",
                     row.names = 1)
expression_data_subset<-read.csv("data/expression_data_for_WGCNA.csv", row.names = 1)
logFC_long_all<-read.csv("data/logFC_tau_all.csv",
                     stringsAsFactors = TRUE)
logFC_long_all$bias_cat <- factor(logFC_long_all$bias_cat, 
                                  levels=c("Unbiased",
                                           "Low",
                                           "Med",
                                           "High",
                                           "Extreme",
                                           "Sex-specific"))
```


## WGCNA

I've subsetted the dataset to only include the genes that showed some form of bias -- otherwise we have >90,000 genes and that's difficult to work with.

```{r}

# make sure we have trait data only for the individuals in the analysis
trait_data<-trait_data[rownames(trait_data) %in% rownames(expression_data_subset),]
```


The first step is to check that the dataset does not contain any outliers or problematic samples (it shouldn't, since we've already cleaned it up for differential expression analysis). Note that WGCNA needs each column to be the genes -- so it's the transposed version of the DESeq dataset.

```{r}
# following the approaches from:
## https://bioinformaticsworkbook.org/tutorials/wgcna.html#gsc.tab=0
## https://cloud.wikis.utexas.edu/wiki/spaces/bioiteam/pages/47714689/Clustering+using+WGCNA
gsg <-goodSamplesGenes(expression_data_subset)
summary(gsg)
gsg$allOK
```

Since everything is ok, we can move on to picking a threshold. 


```{r}
spt <- pickSoftThreshold(expression_data_subset)
```

With this threshold, we want to maximize the $R^2$ value and minimizing mean connectivity.

```{r}
par(mar=c(4,4.5,1,1))
plot(spt$fitIndices[,1],spt$fitIndices[,2],
xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
main = paste("Scale independence"))
text(spt$fitIndices[,1],spt$fitIndices[,2],col="red")
abline(h=0.80,col="red")
```


```{r}
par(mar=c(4,4.5,1,1))
plot(spt$fitIndices[,1], spt$fitIndices[,5],
xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
main = paste("Mean connectivity"))
text(spt$fitIndices[,1], spt$fitIndices[,5], labels= spt$fitIndices[,1],col="red")
```

I think we want to go with 6. 

```{r}
soft_power<-6
# this creates a big matrix
adjacency <- adjacency(expression_data_subset, power = soft_power)
# get the similarity
TOM <- TOMsimilarity(adjacency)
# turn it into dissimilarity
TOM.dissimilarity <- 1-TOM


#creating the dendrogram 
geneTree <- hclust(as.dist(TOM.dissimilarity), 
                   method = "average") 

# identify modules
Modules <- cutreeDynamic(dendro = geneTree, 
                         distM = TOM.dissimilarity, 
                         deepSplit = 2,
                         pamRespectsDendro = FALSE, 
                         minClusterSize = 30)
# assign each module number a color
ModuleColors <- labels2colors(Modules) 
table(ModuleColors)
```


```{r}
# plotting the dendrogram
sizeGrWindow(12,9)

plotDendroAndColors(geneTree, 
                    ModuleColors,"Module",
                    dendroLabels = FALSE, 
                    hang = 0.03,
                    addGuide = TRUE, 
                    guideHang = 0.05,
                    main = "Gene dendrogram and module colors")
```

Now we need to correlate expression modules with traits. 


```{r}
MElist <- moduleEigengenes(expression_data_subset,
                           colors = ModuleColors) 
MEs0 <- MElist$eigengenes 

MEs <- orderMEs(MEs0)
moduleTraitCor <- WGCNA::cor(MEs, trait_data, use= "p")
moduleTraitPvalue <- corPvalueStudent(moduleTraitCor,
                                      nrow(trait_data))
```

```{r}
#Print correlation heatmap between modules and traits
textMatrix<- paste(signif(moduleTraitCor, 2), "\n(", 
						signif(moduleTraitPvalue, 1), ")", sep= "")
dim(textMatrix)<- dim(moduleTraitCor)

# get module counts in correct order
moduleCounts<-table(ModuleColors)
moduleCounts<-moduleCounts[gsub("ME","",dimnames(moduleTraitCor)[[1]])]

# get list of modules significantly associated with sex
sig_modules<-gsub("ME","",
                  rownames(moduleTraitPvalue[signif(moduleTraitPvalue[,"female"], 1)<=0.05,]))
```


```{r, eval=FALSE, fig.width=650, fig.height=550}
par(mar= c(6, 8.5, 3, 3))
#display the correlation values with a heatmap plot

labeledHeatmap(Matrix= moduleTraitCor, 
			xLabels= names(trait_data), 
			yLabels= names(MEs), 
			ySymbols= rep("",nrow(moduleTraitCor)), 
			colorLabels= TRUE, 
			colors= colorRampPalette(c("#af8dc3","#f7f7f7","#7fbf7b"))(50), 
			textMatrix= textMatrix, 
			setStdMargins= FALSE, 
			cex.text= 0.75, 
			zlim= c(-1,1), 
			main="")

```

```{r moduleSexHeatmap, fig.width=350, fig.height=550, eval=FALSE}
#pdf("figs/moduleSexHeatmap.pdf",height = 5.5, width=3.5)
par(mar= c(3, 5, 1, 5))
#display the correlation values with a heatmap plot

labeledHeatmap(Matrix= as.matrix(moduleTraitCor[,1]), 
			xLabels= "", 
			yLabels= names(MEs), 
			ySymbols= paste0("n=",moduleCounts), 
			colorLabels= TRUE, 
			colors= colorRampPalette(c("#af8dc3","#f7f7f7","#7fbf7b"))(50), 
			textMatrix= textMatrix[,1], 
			setStdMargins= FALSE, 
			cex.text= 1, 
			zlim= c(-1,1),
			main="")
text(x=c(1.4,1.4),
     y=c(0.965, 0.05),
     labels = c("female","male"),
     xpd=TRUE)
#dev.off()
```

```{r moduleOrganHeatmap, fig.width=500, fig.height=550, eval=FALSE}
#pdf("figs/moduleOrganHeatmap.pdf", height=5.5, width=5.5)
par(mar= c(3,5,1,3))
#display the correlation values with a heatmap plot

labeledHeatmap(Matrix= moduleTraitCor[,-1], 
			xLabels= names(trait_data[,-1]), 
			yLabels= names(MEs), 
			ySymbols= paste0("n=",moduleCounts), 
			colorLabels= TRUE, 
			colors= colorRampPalette(c("#edf8b1","#7fcdbb","#2c7fb8"))(50), 
			textMatrix= textMatrix[,-1], 
			setStdMargins= FALSE, 
			cex.text= 1, 
			zlim= c(-1,1), 
			main="")
#dev.off()
```

## Compare to other values

### logFC


```{r eval=FALSE}
logFC_long_all$module<-NA


for(col in unique(ModuleColors)){
  genes<-colnames(expression_data_subset)[ModuleColors==col]

  logFC_long_all$module[logFC_long_all$geneID%in%genes]<-col
  
}

logFC_long_all$module<-as.factor(logFC_long_all$module)


write.csv(logFC_long_all,"data/logFC_tau_module_all.csv",row.names = FALSE)
```

```{r}
logFC_long_all<-read.csv("data/logFC_tau_module_all.csv",
                         stringsAsFactors = TRUE)
# sanity checks: missing ones are UB
summary(logFC_long_all$bias[is.na(logFC_long_all$module)])
```


```{r}

liver<-logFC_long_all[logFC_long_all$tissue=="Liver" &
                        !is.na(logFC_long_all$module),]
gill <- logFC_long_all[logFC_long_all$tissue=="Gill" &
                        !is.na(logFC_long_all$module),]
gonad<- logFC_long_all[logFC_long_all$tissue=="Gonad" &
                        !is.na(logFC_long_all$module),]
```

We can make a supplemental figure showing how the genes fall into the various categories per module per tissue.

```{r moduleBiasBarplot, fig.width=1000, fig.height=550}
#pdf("figs/moduleBiasBarplot.pdf", height=5.5, width=10.5)
colors<-rep(levels(as.factor(logFC_long_all$module)), each=3)

par(mfrow=c(1,3), oma=c(1.2,3.4,3,0.5), mar=c(2.5,2,1,0.5))

barplot(table(gill$bias,gill$module),
        col=colors,
        beside=TRUE,
        horiz = TRUE,
        density = c(100,50,0),
        angle=c(45,135,0),
        las=1,
        border=colors,
        xlim=c(0,3000),
        main="gill")


barplot(table(liver$bias,liver$module),
        col=colors,
        beside=TRUE,
        horiz = TRUE,
        density = c(100,50,0),
        angle=c(45,135,0),
        las=1,
        border=colors,
        axisnames = FALSE,
        xlim=c(0,3000),
        main="liver")

barplot(table(gonad$bias,gonad$module),
        col=colors,
        beside=TRUE,
        horiz = TRUE,
        density = c(100,50,0),
        angle=c(45,135,0),
        las=1,
        border=colors,
        axisnames = FALSE,
        xlim=c(0,3000),
        main="gonad")
mtext("number of genes",1,outer=TRUE)
mtext("module",2,outer=TRUE, line=2.1)
outer_legend(
  "top",
  ncol=3,
  c("UB","MB","FB"),
  col="darkgrey",
  density=c(0, 50, 100), 
  angle=c(0,135,45),
  bty="n",
  cex=1.75
)
#dev.off()
```


```{r}
# remove genes not in module analyses
logFC_long_mod<-logFC_long_all[logFC_long_all$geneID %in% colnames(expression_data_subset),]

mod_lm<-lm(rank(logFC_long_mod$logFC)~as.factor(logFC_long_mod$tissue) *as.factor(logFC_long_mod$module))
anova(mod_lm)
mod_labs<-multcompLetters4(aov(mod_lm), 
                           TukeyHSD(aov(mod_lm)))
mod_labs
```
```{r moduleTukeyPlot}
par(mfrow=c(3,1))
plot(TukeyHSD(aov(mod_lm)))
```
```{r}
tukey_tab<-TukeyHSD(aov(mod_lm))
```


```{r}
module_medians<-tapply(logFC_long_mod$logFC,
       paste(logFC_long_mod$tissue, logFC_long_mod$module),
       median,
       na.rm=TRUE)
module_medians[unlist(lapply(sig_modules,
                      grep,
                      x=names(module_medians)))]

lapply(names(table(ModuleColors))[!names(table(ModuleColors)) %in% sig_modules],
       function(col){
         max(module_medians[grep(col, names(module_medians))])
       })
```


```{r, fig.width=10, fig.height=4.5}
pdf("figs/WGCNA_logFC.pdf",height = 4,width = 10)
plot_cols<-unlist(lapply(levels(as.factor(ModuleColors)),rep,times=3))

par(mar=c(4,4,1,1))
plot(x=0:37,
     y=seq(-30,50,length.out=38),
      xaxt='n',
        xlab="",
        ylab="logFC",
     type='n',
     las=1,
     bty='n')
abline(h=0,lty=1,lwd=1)
abline(h=2,lty=2,lwd=2,col="grey")
abline(h=-2,lty=2,lwd=2,col="grey")
vioplot(logFC_long_mod$logFC~logFC_long_mod$tissue + logFC_long_mod$module,
        col=plot_cols,
        add=TRUE,
        bty='n',
        xaxt='n',
        axes=FALSE,
        border="black",
        rectCol=plot_cols,
        lineCol=plot_cols,
        colMed="white",
        plotCentre=TRUE)
grey_vals<-logFC_long_mod[which(logFC_long_mod$module=="grey"),]
grey_vals<-grey_vals$logFC[order(grey_vals$tissue)]
points(x=16:18,
       y=grey_vals,
       col="black",
       bg="grey",
       pch=23,
       cex=1.25)
axis(1,at=1:36,labels=rep(c("Gill","Gonad","Liver"),12),las=2)
axis(2,
     at=c(-30,30),
     c("female-biased","male-biased"),
     tick = 0,
     pos = -2.5)  
     
    

# get the correlation coefficients
cor_vals<-signif(moduleTraitCor, 2)[,1]
names(cor_vals)<-gsub("ME","",names(cor_vals))
cor_vals<-cor_vals[levels(logFC_long_mod$module)]

sig_sex_cor<-which(moduleTraitPvalue[,1]<=0.05)
sig_x<-which(plot_cols %in% gsub("ME","",names(sig_sex_cor)))
sig_x<-sig_x[seq(2,length(sig_x),by=3)]
non_sex_cor<-which(moduleTraitPvalue[,1]>0.05)
non_x<-which(plot_cols %in% gsub("ME","",names(non_sex_cor)))
non_x<-non_x[seq(2,length(non_x),by=3)]
text(x=sig_x,
     y=rep(-30,length(sig_x)),
     labels=cor_vals[names(cor_vals) %in% gsub("ME","",names(sig_sex_cor))],
     font = 2)

dev.off()
```


### Compare to tau

Because tau is a single value per gene, there is no point in looking at it across tissues (unless we wanted to also incorporate logFC, which I do not at this stage). So this is a one-way ANOVA type of analysis.

```{r}
tau_col_dat<-logFC_long_mod[which(logFC_long_mod$module != "grey"),]
tau_col_dat <- tau_col_dat[!is.na(tau_col_dat$module),]
tau_col_dat$module <- factor(tau_col_dat$module)

tau_col_dat <- tau_col_dat[,c("geneID","tau","module")]
tau_col_dat <- unique(tau_col_dat)
```

```{r}
tau_mod<-lm(tau_col_dat$tau~tau_col_dat$module)
summary(tau_mod)
anova(tau_mod)
```


```{r}
par(mfrow=c(2,2))
plot(tau_mod)
```
```{r}
tau_tukey<-TukeyHSD(aov(tau_mod))
tau_mod_lab<-multcompView::multcompLetters4(aov(tau_mod), tau_tukey)
```

```{r, eval=FALSE}
png("figs/tau_modules_vioplot.png",height = 4.5,width=10,units="in",res=300)

plot_cols<-levels(tau_col_dat$module)
par(mar=c(4,4,1,1), oma=c(2,1,1,1))
plot(x=0:length(plot_cols),
     y=seq(0,1,length.out=length(plot_cols)+1),
      xaxt='n',
        xlab="",
        ylab=expression(tau),
     type='n',
     las=1)

vioplot(tau_col_dat$tau~tau_col_dat$module,
        col=plot_cols,
        add=TRUE)

axis(1,
     at=1:length(plot_cols),
     labels=levels(tau_col_dat$module),
     las=2,tick = FALSE)

     
# get the correlation coefficients with sex
cor_vals<-signif(moduleTraitCor, 2)[,1]
names(cor_vals)<-gsub("ME","",names(cor_vals))
cor_vals<-cor_vals[levels(tau_col_dat$module)]

text(x=1:length(plot_cols),
     y=rep(-0.06,length(plot_cols)),
     paste0("r=",cor_vals), 
     xpd=TRUE)

# get labels
vio_labels<-tau_mod_lab[[1]]$Letters
text(x=1:length(plot_cols),
     y=rep(0.95, length(plot_cols)),
     vio_labels)
dev.off()
```

the two most highly correlated modules with male gene expression have significantly higher tissue specificity than the other modules; but the same was not true for the other significantly correlated with males module, nor for the female-correlated module. Strong correlations with up- or down-regulation in tissues was not associated with tissue specificity.

## Identifying hubs

First need to get gene significance and module membership

```{r, eval=FALSE}
# based on https://ngs101.com/how-to-build-gene-co-expression-networks-from-rna-seq-data-using-wgcna-complete-step-by-step-guide-for-absolute-beginners/
# Calculate gene significance (correlation with trait) for sex
gene_trait_cor <- cor(expression_data_subset, 
                      trait_data$female, 
                      use = "pairwise.complete.obs")
gene_trait_pvalue <- corPvalueStudent(as.numeric(gene_trait_cor), nrow(trait_data))
 
# Add gene names to the p-value vector
names(gene_trait_cor) <- colnames(expression_data_subset)
names(gene_trait_pvalue) <- colnames(expression_data_subset)

# create objects to put data into
all_hub_gene_info<-data.frame() 
all_hub_genes<-data.frame()

for(sex_module in sig_modules){

  # Get genes in this module
  module_genes <- colnames(expression_data_subset)[ModuleColors == gsub("ME","",sex_module)]
  cat("Number of genes in", sex_module, "module:", length(module_genes), "\n")
  
   
  # Calculate module membership for these genes
  # MM measures how correlated each gene is with the module eigengene
  gene_module_membership <- cor(expression_data_subset[, module_genes],
                                 MEs[, paste0("ME",sex_module)],
                                 use = "pairwise.complete.obs")
   
  # Create a data frame with gene information
  module_gene_info <- data.frame(
    Gene = module_genes,
    ModuleMembership = as.numeric(gene_module_membership),
    GeneSignificance = gene_trait_cor[module_genes],
    MM_pvalue = corPvalueStudent(as.numeric(gene_module_membership), nrow(trait_data)),
    GS_pvalue = gene_trait_pvalue[module_genes],
    stringsAsFactors = FALSE
  )
   
  # Sort by module membership
  module_gene_info <- module_gene_info[order(-abs(module_gene_info$ModuleMembership)), ]
   
  
  # Calculate connectivity (intramodular connectivity)
  # First, calculate adjacency matrix for genes in this module
  adjacency_matrix <- adjacency(
    expression_data_subset[, module_genes],
    power = soft_power,
    type = "signed"
  )
   
  # Calculate connectivity: sum of connection weights for each gene
  # (subtract 1 to exclude self-connection)
  connectivity <- rowSums(adjacency_matrix) - 1
   
  # Combine all metrics for hub gene identification
  hub_gene_info <- data.frame(
    Gene = module_genes,
    Connectivity = connectivity,
    ModuleMembership = as.numeric(gene_module_membership),
    GeneSignificance =  gene_trait_cor[module_genes],
    stringsAsFactors = FALSE
  )
   
  # Sort by connectivity to identify hubs
  hub_gene_info <- hub_gene_info[order(-hub_gene_info$Connectivity), ]
  hub_gene_info$module <- sex_module 
  
  hub_genes<-hub_gene_info[which(hub_gene_info$Connectivity > quantile(hub_gene_info$Connectivity,0.9) &
                                   hub_gene_info$ModuleMembership > 0.7 &
                                   abs(hub_gene_info$GeneSignificance) > moduleTraitCor[paste0("ME",sex_module),1]),]
  
  # save to bigger data frames
  all_hub_gene_info <- rbind(all_hub_gene_info, hub_gene_info)
  all_hub_genes<-rbind(all_hub_genes, hub_genes)
}

table(all_hub_genes$module)
write.table(all_hub_genes$Gene,"data/hub_gene_IDs.txt",
          row.names=FALSE,
          col.names=FALSE,
          quote=FALSE)
```

      blue       red turquoise 
      220        31       161

```{r, eval=FALSE}
## all_hub_gene_info contains data on all genes 
all_hub_gene_info<-merge(all_hub_gene_info, 
                         logFC_long_all,
                         by.x="Gene",
                         by.y="geneID",
                         all.x=TRUE
                         )
# remove duplicated column
all_hub_gene_info <- all_hub_gene_info[,colnames(all_hub_gene_info) != "module.y"]
colnames(all_hub_gene_info)<-gsub("\\.x","",colnames(all_hub_gene_info))
write.csv(all_hub_gene_info,"data/all_hub_gene_info.csv",row.names = FALSE)
```
```{r, eval=FALSE}
## all_hub_genes contains data ONLY for hub genes in their modules
colnames(all_hub_genes)[5]<-"hub"
all_hub_genes<-merge(all_hub_genes, 
                         logFC_long_mod,
                         by.x="Gene",
                         by.y="geneID",
                         all.x=TRUE
                         )
write.csv(all_hub_genes,"data/all_hub_genes.csv",row.names = FALSE)

```

```{r}
all_hub_gene_info <- read.csv("data/all_hub_gene_info.csv")
all_hub_genes<-read.csv("data/all_hub_genes.csv")
```


#### annotating hubs

Once the hub genes were identified and saved to a file, they were blasted against the _Stigmatopora nigra_ genome, implemented on the RCH. This process is documented in the file `SMA_blast.Rmd`.


### Understanding hub genes

First, let's narrow down the types of genes we might want to annotate: 

- hub genes that are sex-biased (or even just sex-specific?)
- the most highly connected hub genes

#### metrics related to identifying hubs and logFC
```{r ConnectivitySignificanceDist}
par(mfrow=c(1,2), mar=c(4,4,2,2), oma=c(2,2,2,2))
hist(all_hub_gene_info$Connectivity,
     main="",
     xlab="Connectivity",
     ylab="number of hubs")

hist(all_hub_genes$Connectivity,
     col=scales::alpha("cornflowerblue",0.5),
     add=TRUE)

hist(all_hub_gene_info$GeneSignificance,
     main="",
     xlab="Gene significance",
     ylab="") 
hist(all_hub_genes$GeneSignificance,
     col=scales::alpha("cornflowerblue",0.5),
     add=TRUE)
text(x=-0.5, y=-600,"male",xpd=TRUE)
text(x=0.5, y=-600,"female",xpd=TRUE)
# note: some genes are double-counted if they appeared in multiple tissues

```

is there a way to better understand module-sexbias relationships?

let's go module by module, maybe? and focus on highly significant genes in wgcna

```{r}
organ_cols<-c("Gill" = "#20B2AA",
              "Gonad" = "#EEB422", 
              "Liver" = "#EE8262")
organ_shapes <- c(Gill = 16, Gonad = 15,Liver = 17)
```


```{r ModuleHubGeneSignificance}
par(mfrow=c(1,3), mar=c(2,2,2,2))
for(module in sig_modules){
  this_mod<-all_hub_genes[all_hub_genes$module==module,]
  
  plot(this_mod$GeneSignificance, 
       this_mod$logFC,
       col=scales::alpha(organ_cols[this_mod$tissue],0.5),
       bty='l',
       pch=organ_shapes[this_mod$tissue],
       main=module,
       xlab="",
       ylab="")
}
mtext("Gene Significance",1,outer=TRUE)
mtext("logFC",2,outer=TRUE)
outer_legend("top",
             names(organ_cols),
             col=organ_cols,
             pch=organ_shapes,
             bty='n',
             ncol=3, 
             cex=1.5)

```

```{r ModuleHubConnectivity}
par(mfrow=c(1,3), mar=c(2,2,2,2))
for(module in sig_modules){
  this_mod<-all_hub_genes[all_hub_genes$module==module,]
  
  plot(this_mod$Connectivity, 
       this_mod$logFC,
       col=scales::alpha(organ_cols[this_mod$tissue],0.5),
       bty='l',
       pch=organ_shapes[this_mod$tissue],
       main=module,
       xlab="",
       ylab="")
}
mtext("Connectivity",1,outer=TRUE)
mtext("logFC",2,outer=TRUE)
outer_legend("top",
             names(organ_cols),
             col=organ_cols,
             pch=organ_shapes,
             bty='n',
             ncol=3, 
             cex=1.5)

```
```{r ModuleHubModuleMembership}
par(mfrow=c(1,3), mar=c(2,2,2,2))
for(module in sig_modules){
  this_mod<-all_hub_genes[all_hub_genes$module==module,]
  
  plot(this_mod$ModuleMembership, 
       this_mod$logFC,
       col=scales::alpha(organ_cols[this_mod$tissue],0.5),
       bty='l',
       pch=organ_shapes[this_mod$tissue],
       main=module,
       xlab="",
       ylab="")
}
mtext("Module Membership",1,outer=TRUE)
mtext("logFC",2,outer=TRUE)
outer_legend("top",
             names(organ_cols),
             col=organ_cols,
             pch=organ_shapes,
             bty='n',
             ncol=3, 
             cex=1.5)

```

#### Exploring specific expression patterns


```{r}
table(paste0(all_hub_genes$bias, all_hub_genes$tissue),
      all_hub_genes$hub)
```

```{r}
odd_blue<-all_hub_genes$Gene[which(all_hub_genes$hub=="blue" & 
                all_hub_genes$tissue=="Gonad" &
                all_hub_genes$bias!="MB")]
all_hub_genes[all_hub_genes$Gene==odd_blue,]
```
```{r}
blue_bias_counts<-tapply(all_hub_genes$bias[all_hub_genes$hub=="blue"],
                         all_hub_genes$Gene[all_hub_genes$hub=="blue"],
                    function(x){
                      return(paste(x,collapse="_"))
                    })
oddities<-blue_bias_counts[!blue_bias_counts%in% c("UB_MB_UB","UB_UB_MB","MB_UB_UB")]
all_hub_genes[all_hub_genes$Gene %in% names(oddities),]
```

```{r}
hub_blast_ann<-read.csv("data/hub_blast_ann.csv")
unique(hub_blast_ann[hub_blast_ann$qseqid %in% names(oddities),
                     c("product","GO","Symbol","Gene_Name","TERM","ONTOLOGY","DEFINITION")])
```
```{r}
red_hubs<-hub_blast_ann[hub_blast_ann$hub=="red",]
matches<-tapply(red_hubs$product,red_hubs$qseqid,unique)
 length(matches[!is.na(matches)])
```

```{r}
turquoise<-hub_blast_ann[hub_blast_ann$hub=="turquoise",]
matches<-tapply(turquoise$product,turquoise$qseqid,unique)
 length(matches[!is.na(matches)])
```
```


