---
title: "Annotating genes based on S. nigra genome"
author: "Sarah P. Flanagan"
output:
  bookdown::pdf_document2:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    toc: true
header-includes: >
  \usepackage{lipsum}
  \usepackage{float}
  \floatplacement{figure}{H}
editor_options:
  chunk_output_type: console
always_allow_html: true
---

This document contains some code that was adapted from other scripts, including code originally authored by Coley Tosto and Emily Beasley.

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir='../',fig_path="../imgs/")
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```
```{r setdir}
knitr::opts_knit$set(root.dir='../',
                     fig_path="../figs/")
```
```{r}
logFC_tau_all<-read.csv("data/logFC_tau_all.csv")
logFC_long_all$bias_cat <- factor(logFC_long_all$bias_cat, 
                                  levels=c("Unbiased",
                                           "Low",
                                           "Med",
                                           "High",
                                           "Extreme",
                                           "Sex-specific"))

```


```{r}
filter_blast<-function(data_frame){
 
  #Pull out the Unique Trinity gene IDs
  uniqueID <- unique(data_frame[,1])
  
  #Create an empty dataframe to store intermediate results in
  output <- data.frame(matrix(data = NA, ncol = ncol(data_frame), nrow = 0))
  colnames(output) <-  colnames(data_frame)
  
  #Generate a for loop that pulls out the lowest e-value + highest % identity for each gene
  for(gene in uniqueID){
    
    #Subset the dataset for each Trinity gene
    this_trin_gene<- data_frame[which(data_frame[,1] == gene),]
    #Pull out gene with smallest e-value
    uni_gene_subset_lowe <- this_trin_gene[which(this_trin_gene$evalue == min(this_trin_gene$evalue)),]
    #In case mult. genes have same e-value, pull out gene with highest % identity
    uni_gene_subset_lowe_highpid <- uni_gene_subset_lowe[which(uni_gene_subset_lowe$pident == max(uni_gene_subset_lowe$pident)),]
    
    # keep only one of multiple identical rows
    uni_gene_subset_lowe_highpid <- unique(uni_gene_subset_lowe_highpid)
    
    #Add the final gene into the empty dataframe from above
    output <- rbind(output, uni_gene_subset_lowe_highpid)
  }
  
  return(output)
}
```


```{r}
parse_gff<-function(filename){
  gff <- read.delim(filename,
                  comment.char = "#", 
                  header = FALSE)
  colnames(gff) <- c("chromosome","method","type","start","stop","x","strand","y","description")
  return(gff)
}
```

```{r}
blast2gff<-function(sid,gff){
  this_gff<-gff[grep(sid, gff$description),]
  product<-paste(unique(gsub("^.*product=(.*)\\;protein_id.*$","\\1",this_gff$description)),collapse=";")
  gene_id<-paste(unique(gsub("^.*GeneID\\:(\\d+)\\,.*$","\\1",this_gff$description)),collapse=";")
  if(length(product)<1){ browser() }
  return(data.frame(product=product, GeneID=gene_id))
}
```
```{r}
get_condensed_annotations<-function(this_gene, annotations){
  
  this_ann<-annotations[annotations$GeneID==as.numeric(this_gene["GeneID"]),]
  
  return(paste(paste(this_ann$Qualifier, this_ann$TERM),collapse=";"))
  
}

```
```{r}
get_all_annotations<-function(this_gene, annotations){
  #browser()
  this_ann<-annotations[annotations$GeneID==as.numeric(this_gene["GeneID"]),]
  
  return(this_ann)
  
}
```



I have three categories of genes that I've annotated:

- sex-biased genes
- highly sex-biased and also tissue-specific genes
- hub genes from the WGCNA


They all followed the same steps so I will put them in this document. The identification and export of the gene IDs happened in each of the relevant files.

# Running diamond blast on RCH


```{bash, eval=FALSE}
module load DIAMOND/2.1.11-GCC-13.3.0
diamond makedb --in stigmatopora_nigra/data/GCF_051989575.1/protein.faa --db s_nigra
```

```{bash, eval=FALSE}
#!/bin/bash

#
#SBATCH --job-name=diamondblastx
#
#SBATCH --ntasks=1
#SBATCH --time=25:00:00
#SBATCH --mem-per-cpu=3GB
#SBATCH --cpus-per-task=8
#SBATCH --error=job_%x_%j.err
#SBATCH --output=job_%x_%j.out


module load prefix/latest
module load DIAMOND/2.1.11-GCC-13.3.0

cd /home/spf50/S_macropterygia/

# process highly tissue specific and sex biased genes
./subset_fasta_file -c -f SMA_supertranscriptome.fasta -l sex_biased_trinity_genes.txt
mv SMA_supertranscriptome..subset.fasta sex_biased_genes.fasta

# process highly tissue specific and sex biased genes
./subset_fasta_file -c -f SMA_supertranscriptome.fasta -l tissue_sex_biased_IDs.txt
mv SMA_supertranscriptome..subset.fasta tissue_sex_biased.fasta

# process hub genes
./subset_fasta_file -c -f SMA_supertranscriptome.fasta -l hub_gene_IDs.txt
mv SMA_supertranscriptome..subset.fasta hub_genes.fasta

# run diamond blastx
diamond blastx --threads 8 --db s_nigra --out sex_biased_genes.dmdblastx --query sex_biased_genes.fasta --outfmt 6

diamond blastx --threads 8 --db s_nigra --out tissue_sex_biased.dmdblastx --query tissue_sex_biased.fasta --outfmt 6

diamond blastx --threads 8 --db s_nigra --out hub_genes.dmdblastx --query hub_genes.fasta --outfmt 6


```

## Preparing S. nigra info 


I need to merge the annotations in the gff with GO categories from GO.db.

```{r, eval=FALSE}
annotations<-read.delim("stigmatopora_nigra/GCF_051989575.1-RS_2025_11_gene_ontology.gaf",
                        comment.char="!", 
                        header = FALSE)
colnames(annotations)<-c(
  "DB","GeneID","Symbol","Qualifier","GO_ID","Reference","Evidence_Code","With_From","Aspect","Gene_Name","Gene_Synonym","Type","Taxon","Date","Assigned_By","Annot_Ext","Gene_Product_Form_ID"
)
library(GO.db)
# get keys for selection
go <- keys(GO.db, keytype="GOID")
goids <- select(GO.db,
                columns=c("GOID","TERM","ONTOLOGY","DEFINITION"),
                keys=go, 
                keytype="GOID")
annotations<-merge(annotations,
                   goids,
                   by.x="GO_ID",
                   by.y="GOID",
                   all.x=TRUE)
write.csv(annotations,
          "stigmatopora_nigra/go_annotations.csv",
          row.names=FALSE)
```
```{r}
gff<-parse_gff("stigmatopora_nigra/genomic.gff")
annotations<-read.csv("stigmatopora_nigra/go_annotations.csv")
```



# Filtering, matching and parsing 

## Sex-biased genes


### Filtering Blast results

Because of the way BLAST was run in the RCH, we can have multiple hits for one gene, I want to filter that out so that there is only one hit kept for each gene.



```{r filterBlast, eval=FALSE}
blast_output_all <- read.delim("data/sex_biased_genes.dmdblastx",
                               header=FALSE)
colnames(blast_output_all) <- c(
  "qseqid",
  "sseqid",
  "pident",
  "length",
  "mismatch", 
  "gapopen", 
  "qstart", 
  "qend", 
  "sstart", 
  "send",
  "evalue", 
  "bitscore")

blast_filtered<-filter_blast(blast_output_all)
# note: this still contains some repeats -- same criteria but different sseqids

write.csv(blast_filtered,"sex_bias_blast_filtered.csv")
```


```{r blastAnnotate, eval=FALSE}
# run on RCH
blast_filtered <- read.csv("data/sex_bias_blast_filtered.csv", row.names = 1)
gff<-parse_gff("stigmatopora_nigra/genomic.gff")
annotations<-read.csv("stigmatopora_nigra/go_annotations.csv")

blast_info<-as.data.frame(do.call(rbind, lapply(blast_filtered$sseqid, 
                                                blast2gff,gff=gff)))

blast_filtered$GeneID<-blast_info$GeneID
blast_filtered$product<-blast_info$product

# get the GO annotations
go_anns<-apply(blast_filtered, 1, 
               get_condensed_annotations, annotations=annotations)
blast_filtered$GO<-go_anns
write.csv(blast_filtered,
          "data/sex_bias_blast_filtered_annotated.csv",
          row.names=FALSE)
```




## Sex-biased AND tissue specific

(there should be some overlap between the previous set of annotations and the current set)


```{r}
# read in the blast data
tissue_sex_blast<-read.table("data/tissue_sex_biased.dmdblastx",header=FALSE)
colnames(tissue_sex_blast)<-c("qseqid", "sseqid", "pident", "length", "mismatch", "gapopen", "qstart", "qend", "sstart", "send", "evalue", "bitscore")
```

We can summarise the matches before filtering:

```{r}
matched_genes<-unique(tissue_sex_blast$qseqid)

length(matched_genes)
#[1] 851
unmatched<-unique(tissue_sex_biased$geneID [!tissue_sex_biased$geneID  %in% matched_genes])

length(unmatched)
#498
```

Now we can filter and annotate.

```{r, eval=FALSE}
tissue_sex_blast_filtered<-filter_blast(tissue_sex_blast)

tissue_sex_blast_info<-as.data.frame(do.call(rbind,
                                        lapply(tissue_sex_blast_filtered$sseqid, 
                                                    blast2gff,gff=gff)))
tissue_sex_blast_filtered$GeneID<-tissue_sex_blast_info$GeneID
hub_blast_filtered$product<-tissue_sex_blast_info$product


write.csv(tissue_sex_blast_filtered,"data/tissue_sex_blast_filtered.csv",row.names = FALSE)
```
```{r}
tissue_sex_blast_filtered<-read.csv("data/tissue_sex_blast_filtered.csv")
# get condensed versions
tissue_sex_blast_filtered$GO<-apply(tissue_sex_blast_filtered,
                                    1, 
                                    get_condensed_annotations,
                                    annotations=annotations)
```


```{r}
# look at how many annotations there were per gene
all_go_anns<-do.call(rbind,apply(tissue_sex_blast_filtered, 1, 
                                 get_all_annotations, annotations=annotations))


tissue_sex_blast_ann<-merge(tissue_sex_blast_filtered,
                            all_go_anns, 
                            by.x="GeneID",
                            by.y="GeneID", 
                            all.x=TRUE, 
                            all.y=FALSE)
```

```{r plotNumGOperGene}
all_go_anns_tab<-table(all_go_anns$TERM)
hist(all_go_anns_tab,
     main="",
     xlab="number of GO annotations per gene",
     ylab="number of genes")
```

```{r}
tau_cutoff<-round(quantile(logFC_tau_all$tau,
                           probs = 0.9,
                           na.rm=TRUE),2)
logFC_cutoff<-round(quantile(abs(logFC_tau_all$logFC),probs = 0.9, na.rm=TRUE),2)
tissue_sex_biased<-logFC_tau_all[which(logFC_tau_all$tau > tau_cutoff &
                       abs(logFC_tau_all$logFC) > logFC_cutoff &
                         logFC_tau_all$bias!="UB"),]
```


```{r}
tissue_sex_blast_ann<-merge(tissue_sex_blast_ann, 
                            tissue_sex_biased,
                            by.x="qseqid",
                            by.y="geneID",
                            all.x=TRUE, 
                            all.y=TRUE)
write.csv(tissue_sex_blast_ann,"data/tissue_sex_blast_ann.csv", row.names = FALSE,quote=FALSE)
```

```{r}
liver_MB_table<-as.data.frame(table(tissue_sex_blast_ann$TERM[tissue_sex_blast_ann$bias=="MB" &
                                                   tissue_sex_blast_ann$tissue=="Liver"]))
liver_MB_table<-liver_MB_table[liver_MB_table$Freq>0,]
liver_MB_table$bias<-"MB"
liver_MB_table$tissue<-"Liver"


gill_MB_table<-as.data.frame(table(tissue_sex_blast_ann$TERM[tissue_sex_blast_ann$bias=="MB" &
                                                   tissue_sex_blast_ann$tissue=="Gill"]))
gill_MB_table<-gill_MB_table[gill_MB_table$Freq>0,]
gill_MB_table$bias<-"MB"
gill_MB_table$tissue<-"Gill"


gonad_MB_table<-as.data.frame(table(tissue_sex_blast_ann$TERM[tissue_sex_blast_ann$bias=="MB" &
                                                   tissue_sex_blast_ann$tissue=="Gonad"]))
gonad_MB_table<-gonad_MB_table[gonad_MB_table$Freq>0,]
gonad_MB_table$bias<-"MB"
gonad_MB_table$tissue<-"Gonad"

liver_FB_table<-as.data.frame(table(tissue_sex_blast_ann$TERM[tissue_sex_blast_ann$bias=="FB" &
                                                   tissue_sex_blast_ann$tissue=="Liver"]))
liver_FB_table<-liver_FB_table[liver_FB_table$Freq>0,]
liver_FB_table$bias<-"FB"
liver_FB_table$tissue<-"Liver"


gill_FB_table<-as.data.frame(table(tissue_sex_blast_ann$TERM[tissue_sex_blast_ann$bias=="FB" &
                                                   tissue_sex_blast_ann$tissue=="Gill"]))
gill_FB_table<-gill_FB_table[gill_FB_table$Freq>0,]
gill_FB_table$bias<-"FB"
gill_FB_table$tissue<-"Gill"

gonad_FB_table<-as.data.frame(table(tissue_sex_blast_ann$TERM[tissue_sex_blast_ann$bias=="FB" &
                                                   tissue_sex_blast_ann$tissue=="Gonad"]))
gonad_FB_table<-gonad_FB_table[gonad_FB_table$Freq>0,]
gonad_FB_table$bias<-"FB"
gonad_FB_table$tissue<-"Gonad"

tissue_sex_go_table<-rbind(
  gonad_MB_table,
  liver_MB_table,
  gill_MB_table,
  gonad_FB_table,
  liver_FB_table,
  gill_FB_table
)
colnames(tissue_sex_go_table)<-c("GO_Term","Number of matches", "bias","tissue")
write.csv(tissue_sex_go_table, "data/tissue_sex_GO_table.csv",row.names = FALSE, quote=FALSE)
```

## Hub genes


```{r}
# read in the blast data
hub_blast<-read.table("data/hub_genes.dmdblastx",header = FALSE)
colnames(hub_blast)<-c("qseqid", "sseqid", "pident", "length", "mismatch", "gapopen", "qstart", "qend", "sstart", "send", "evalue", "bitscore")
```


```{r}
matched_genes<-unique(hub_blast$qseqid)

length(matched_genes)
#[1] 320
unmatched<-unique(all_hub_genes$Gene[!all_hub_genes$Gene %in% matched_genes])
length(unmatched)
#139
```


```{r, eval=FALSE}
hub_blast_filtered<-filter_blast(hub_blast)

hub_blast_info<-as.data.frame(do.call(rbind, lapply(hub_blast_filtered$sseqid, 
                                                    blast2gff,gff=gff)))
hub_blast_filtered$GeneID<-hub_blast_info$GeneID
hub_blast_filtered$product<-hub_blast_info$product


write.csv(hub_blast_filtered,"data/hub_genes_blast_filtered.csv",row.names = FALSE)
```

```{r}
hub_blast_filtered<-read.csv("data/hub_genes_blast_filtered.csv")
hub_blast_filtered$GO<-apply(hub_blast_filtered, 1, 
                             get_condensed_annotations, annotations=annotations)
```



```{r}
all_go_anns<-do.call(rbind,apply(hub_blast_filtered, 1, 
                                 get_all_annotations, annotations=annotations))

hub_blast_ann<-merge(hub_blast_filtered,
                            all_go_anns, 
                            by.x="GeneID",
                            by.y="GeneID", 
                            all.x=TRUE, 
                            all.y=FALSE)
```
```{r}
hub_blast_ann<-merge(hub_blast_ann, 
                            all_hub_gene_info,
                            by.x="qseqid",
                            by.y="Gene",
                            all.x=TRUE, 
                            all.y=TRUE)
```

```{r}
all_go_anns_tab<-table(all_go_anns$TERM)
hist(all_go_anns_tab)

```


```{r}
liver_MB_table<-as.data.frame(table(hub_blast_ann$TERM[hub_blast_ann$bias=="MB" &
                                           hub_blast_ann$tissue=="Liver"],
                      hub_blast_ann$module[hub_blast_ann$bias=="MB" &
                                           hub_blast_ann$tissue=="Liver"]))
liver_MB_table<-liver_MB_table[liver_MB_table$Freq>0,]
liver_MB_table$bias<-"MB"
liver_MB_table$tissue<-"Liver"

gill_MB_table<-as.data.frame(table(hub_blast_ann$TERM[hub_blast_ann$bias=="MB" &
                                           hub_blast_ann$tissue=="Gill"],
                      hub_blast_ann$module[hub_blast_ann$bias=="MB" &
                                           hub_blast_ann$tissue=="Gill"]))
gill_MB_table<-gill_MB_table[gill_MB_table$Freq>0,]
gill_MB_table$bias<-"MB"
gill_MB_table$tissue<-"Gill"

gonad_MB_table<-as.data.frame(table(hub_blast_ann$TERM[hub_blast_ann$bias=="MB" &
                                           hub_blast_ann$tissue=="Gonad"],
                      hub_blast_ann$module[hub_blast_ann$bias=="MB" &
                                           hub_blast_ann$tissue=="Gonad"]))
gonad_MB_table<-gonad_MB_table[gonad_MB_table$Freq>0,]
gonad_MB_table$bias<-"MB"
gonad_MB_table$tissue<-"Gonad"

liver_FB_table<-as.data.frame(table(hub_blast_ann$TERM[hub_blast_ann$bias=="FB" &
                                           hub_blast_ann$tissue=="Liver"],
                      hub_blast_ann$module[hub_blast_ann$bias=="FB" &
                                           hub_blast_ann$tissue=="Liver"]))
liver_FB_table<-liver_FB_table[liver_FB_table$Freq>0,]
liver_FB_table$bias<-"FB"
liver_FB_table$tissue<-"Liver"

gill_FB_table<-as.data.frame(table(hub_blast_ann$TERM[hub_blast_ann$bias=="FB" &
                                           hub_blast_ann$tissue=="Gill"],
                      hub_blast_ann$module[hub_blast_ann$bias=="FB" &
                                           hub_blast_ann$tissue=="Gill"]))
# there were no matches
#gill_FB_table<-gill_FB_table[gill_FB_table$Freq>0,]
#gill_FB_table$bias<-"FB"
#gill_FB_table$tissue<-"Gill"

gonad_FB_table<-as.data.frame(table(hub_blast_ann$TERM[hub_blast_ann$bias=="FB" &
                                           hub_blast_ann$tissue=="Gonad"],
                      hub_blast_ann$module[hub_blast_ann$bias=="FB" &
                                           hub_blast_ann$tissue=="Gonad"]))
gonad_FB_table<-gonad_FB_table[gonad_FB_table$Freq>0,]
gonad_FB_table$bias<-"FB"
gonad_FB_table$tissue<-"Gonad"

module_go_table<-rbind(
  gonad_MB_table,
  liver_MB_table,
  gill_MB_table,
  gonad_FB_table,
  liver_FB_table
)
colnames(module_go_table)<-c("GO_Term","module","Number of matches", "bias","tissue")
write.csv(module_go_table, "data/module_GO_table.csv",row.names = FALSE, quote=FALSE)
```
