---
title: "Differential Expression Analysis in _Stigmatopora argus_"
output:
  bookdown::pdf_document2:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    toc: true
header-includes: >
  \usepackage{lipsum}
  \usepackage{float}
  \floatplacement{figure}{H}
editor_options:
  chunk_output_type: console
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir='../',fig_path="../imgs/")
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```

This .Rmd is the analysis that follows the reassignment of identity to _S. argus_ samples after initial visualisation and inspection of previous sample identities was thought to be incorrect.

``` {r library, include = TRUE, message = FALSE, warning = FALSE}
setwd("~/Documents/GitHub/pipefish_sbge/stigmatopora/S_argus")
#This is a cohesive list of all the libraries used in this document
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
library(PCAtools)
library(ggpubr)
library(cowplot)
library(UpSetR)
library(ggplot2)
library(kableExtra)
library(tidyverse)
library(knitr)
library(magick)
library(multcompView)
library(spfTools)
```

``` {r read-data}
#The abundance matrix generated via salmon and tximport to be used for the DE analysis
txi.argus <- readRDS("data/new.txi.salmon_SA.RDS")

#The samples file generated for tximport
samples <- read.table("data/SA_sample_ID_new.txt", header = TRUE)

#Make sure the conditions are in the samples file as a factor
samples$sex <- as.factor(samples$sex)
samples$organ <- as.factor(samples$organ)

```

The package `DESeq2` was used for the differential expression analysis outlined below.

# Single factor analysis - Comparing Males v Females across all organs
To analyse your data with DESeq2 you must first generate the DESeqDataSet. In order to do this we need the abundance matrix generated with `tximport` and a `samples` file that lays out all of the conditions. The model for this single factor analysis was run as counts ~ Sex.

```{r DESeqDataSet, message=FALSE, warning=FALSE}
#Create the DESeq dataset
dds_SA <- DESeqDataSetFromTximport(txi.argus, 
                                   colData = samples,
                                   design = ~ sex)

```

The data is then pre-filtered to remove low gene counts before running further DESeq2 functions. By doing this we remove rows in which there are very few reads thus reducing the memory size of the `dds` object and increasing the speed at which we can use the transformation and testing functions in DESeq2.

The cutoff here was to remove rows that had counts fewer than 10 across all samples.

```{r pre-filtering, message=FALSE, warning=FALSE}
#only keeping rows that have at lead 10 reads total
keep <- rowSums(counts(dds_SA)) >= 10
dds_SA <- dds_SA[keep, ]

```

After filtering we can now perform the standard differential expression analysis that is wrapped into DESeq2.

```{r diff-exp, message=FALSE, warning=FALSE}
#Generate the expression values
dds_SA_exp <- DESeq(dds_SA)

#Compile the results
res <- results(dds_SA_exp)
res
```

Once that has finished we can now start exploring some of the single-factor analysis results
```{r investigate-DESeq-results-SA}
##Ordering our results based on p-value
resOrdered <- res[order(res$pvalue),]
resOrdered

summary(res)

#How many ADJUSTED p-values were less than 0.1?
sum(res$padj < 0.1, na.rm = TRUE)

#Looking at  an alpha=0.05, the default is 0.1
res05 <- results(dds_SA_exp, alpha = 0.05)
summary(res05)
head(res05)
sum(res05$padj < 0.05, na.rm = TRUE)

```

## Visualizing the results
### MA-plot - MvF Differntial expression
Generate an MA-plot to show the log2 fold changes attributable to sex over the mean of normalized counts for all of the samples in the `dds`. Points will be colored if the adjusted p-value is less that 0.1.

```{r MA-plot, echo=FALSE, fig.cap="LogFC versus the mean normalized count for all of the genes. Points that are blue have a p-value less than 0.1. The two plots are the same with the only difference coming from an adjusted y-limit. Points that are triangles represent genes with a fold change higher/lower than the y-limit."}

par(mfrow=c(1,2))
plotMA(res, ylim = c(-2,2))
plotMA(res)

```

### Heatmap - overall expression
We can also generate a heat map to look at overall expression levels across our samples. Note, this is not differentially expressed genes.

```{r heatmap, echo=FALSE, fig.cap= "Heatmap showing the expression level across all organs for the top 20 genes with the highest expression levels."}
df <- as.data.frame(samples[,c("sex", "organ")])
rownames(df) <- samples$ID
df$organ <- as.character(df$organ)

organ_cols<-c("gill" = "#20B2AA",
              "gonad" = "#EEB422", 
              "liver" = "#EE8262")

sex_cols<-c("F" = "#7fc97f", "M" = "#beaed4")

ann_colors = list(sex=sex_cols,
                  organ=organ_cols)

col_order <- c(rownames(df[df$sex=="F" & df$organ=="gill",]),
               rownames(df[df$sex=="M" & df$organ=="gill",]),
               rownames(df[df$sex=="F" & df$organ=="gonad",]),
               rownames(df[df$sex=="M" & df$organ=="gonad",]),
               rownames(df[df$sex=="F" & df$organ=="liver",]),
               rownames(df[df$sex=="M" & df$organ=="liver",]))


#Pull out the top 20 rows with the most expression
select <- order(rowMeans(counts(dds_SA_exp, normalized = TRUE)),
                decreasing = TRUE)[1:20]


#Transform the data
vsd <- vst(dds_SA_exp, blind=FALSE)

#Run the heat map with the function pheatmap
pheatmap(assay(vsd)[select, col_order], 
         cluster_rows = FALSE, 
         show_rownames = TRUE, 
         cluster_cols = FALSE, 
         annotation_col = df, 
         annotation_colors = ann_colors)
```

From the heatmap (Fig. \@ref(fig:heatmap)) we can see that for all of these top 20 expressed genes the highest expression in found in the gonads (specifically, three ovaries and two testes), which is different from the other two _Stigmatopora_ spp. and all of the _Syngnathus_ species that had the highest expression in the liver. There are also a few genes that appear to be highly expressed across all of the tissue types. 

With this heatmap we can also pull out the names of the Trinity genes that are showing this high expression and BLAST the corresponding sequences to see what the genes are.

```{r heatmap_genes}
#Pull out the corresponding trinity_geneIDS that are plotted in the heatmap
heatmap_TG <- cbind(row.names(assay(vsd)[select,]))
#Run once to create the file and then commented out
#write.table(heatmap_TG,
 #           'heatmap_trinitygenes.txt',
  #          sep = "",
   #         quote=FALSE,
    #        row.names = FALSE,
     #       col.names = FALSE)
```

### Sample-dist Heatmap

We can then generate a sample-to-sample distances heatmap that will give an overview of the similarities and differences between our samples. The darker the colour, the more similar the samples are. The diagonal line of very dark blue represents the comparisson between the same samples.

```{r sample-dist, echo=FALSE, fig.cap="Sample-to-sample distances heatmap showing the similarities and differences between all of the samples. The darker the colour, the more similar the samples are. The diagonal line of very dark blue represents the comparisons between the same samples.", warning=FALSE}
#Calculate the sample-to-sample distances from the transformed count matrix
sampleDists <- dist(t(assay(vsd)))

#Transform into a matrix
sampleDistsMatrix <- as.matrix(sampleDists)
rownames(sampleDistsMatrix) <- paste(vsd$sex, vsd$organ, sep = "-")
colnames(sampleDistsMatrix) <- vsd$ID

#Run the heatmap
colors <- colorRampPalette(rev(brewer.pal(9, 'Blues')))(255)
pheatmap(sampleDistsMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists, 
         col = colors)
```

This is a definite improvement on the original sample distribution heatmap, although we still have one female gill (S39) clustering with livers, one male gonad clustering with gills (S46), and one female liver clustering with gonads (S32), so we might need to sort that out. We'll see what the PCA plots look like first. 

### PCA plots

```{r pca-pairs-plot, echo=FALSE, fig.cap="Principal components analysis pairsplot showing PCA axis 1 - 4."}
#Generate the PCA data to create the pairs plot, using the PCAtools package
p <- pca(assay(vsd), 
         metadata = colData(dds_SA))

#Create the pairs plot
pairsplot(p,
          components = getComponents(p, c(1:4)),
          colby = "organ",
          triangle = FALSE,
          hline = 0, vline = 0,
          colkey = c("gill" = "#6E8B3D75", 
                     "gonad" = "#EEB42275", 
                     "liver" = "#EE826275"),
          shape = "sex",
          shapekey = c(16, 17),
          pointSize = 3,
          margingaps = unit(c(0.1, 0.1, 0.1, 0.1), 'cm'),
          legendPosition = "left", legendIconSize = 3, legendLabSize = 10)

```

For the most part, samples are clustering with like samples (gills with gills, etc.). S39 still appears to be an outlier, and the S32 is clustering outside of any of the organ groups. 

Next we can look at PCA plots for our samples to see how our different samples are grouping.
```{r pca-plot, echo=FALSE, message=FALSE, warning=FALSE}

#Generate the PCA dataframe
pca <- prcomp(t(assay(vsd)))
pca_plotting_data<-as.data.frame(pca$x)

#Add the metadata to the PCA dataset
pca_plotting_data$organ <- samples$organ
pca_plotting_data$sex <- samples$sex
pca_plotting_data$ID <- samples$ID

#Calculate the percent variation attributed to each axis 
percentVar <- round(p$variance, digits = 2)


my_colors <- c("gill" = "#6E8B3D75", "gonad" = "#EEB42275", "liver" = "#EE826275")
my_shapes <- c(16, 17)

ggplot(pca_plotting_data, aes(x=PC1, y=PC2, color = organ, shape = sex)) +
  scale_color_manual(values = my_colors) +
  scale_shape_manual(values = my_shapes) +
  geom_point(size = 5) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) +
  geom_hline(yintercept = 0,
             linetype = 'dashed') +
  geom_vline(xintercept = 0,
             linetype = 'dashed') + 
  geom_text(aes(label = ID), vjust = -1, hjust = 0.5)
```


### Let's get rid of S32 and S39
Based on the heatmaps and PCA plots, I am going to remove S32 and S39.

```{r rm-samples-DESeqDataSet, message=FALSE, warning=FALSE}
#Create the DESeq dataset
dds_SA <- DESeqDataSetFromTximport(txi.argus, 
                                   colData = samples,
                                   design = ~ sex)

```


```{r rm-samples-pre-filtering, message=FALSE, warning=FALSE}
#only keeping rows that have at lead 10 reads total
keep <- rowSums(counts(dds_SA)) >= 10
dds_SA <- dds_SA[keep, ]

#remove S32, S39, and S46
dds_SA <- dds_SA[, !(dds_SA$ID %in% c("S32", "S39", "S46"))]

```

After filtering we can now perform the standard differential expression analysis that is wrapped into DESeq2.

```{r rm-samples-diff-exp, message=FALSE, warning=FALSE}
#Generate the expression values
dds_SA_exp <- DESeq(dds_SA)

#Compile the results
res <- results(dds_SA_exp)
res
```


### Heatmap - overall expression
Let's visualise again - generate a heat map to look at overall expression levels across our samples. Note, this is not differentially expressed genes.

```{r rm-samples-heatmap, echo=FALSE, fig.cap= "Heatmap showing the expression level across all organs for the top 20 genes with the highest expression levels."}
samples <- samples[!(samples$ID %in% c("S32", "S39","S46")),]
df <- as.data.frame(samples[,c("sex", "organ")])
rownames(df) <- samples$ID
df$organ <- as.character(df$organ)

organ_cols<-c("gill" = "#20B2AA",
              "gonad" = "#EEB422", 
              "liver" = "#EE8262")

sex_cols<-c("F" = "#7fc97f", "M" = "#beaed4")

ann_colors = list(sex=sex_cols,
                  organ=organ_cols)

col_order <- c(rownames(df[df$sex=="F" & df$organ=="gill",]),
               rownames(df[df$sex=="M" & df$organ=="gill",]),
               rownames(df[df$sex=="F" & df$organ=="gonad",]),
               rownames(df[df$sex=="M" & df$organ=="gonad",]),
               rownames(df[df$sex=="F" & df$organ=="liver",]),
               rownames(df[df$sex=="M" & df$organ=="liver",]))


#Pull out the top 20 rows with the most expression
select <- order(rowMeans(counts(dds_SA_exp, normalized = TRUE)),
                decreasing = TRUE)[1:20]


#Transform the data
vsd <- vst(dds_SA_exp, blind=FALSE)

#Run the heat map with the function pheatmap
pheatmap(assay(vsd)[select, col_order], 
         cluster_rows = FALSE, 
         show_rownames = TRUE, 
         cluster_cols = FALSE, 
         annotation_col = df, 
         annotation_colors = ann_colors)
```

From the heatmap (Fig. \@ref(fig:heatmap)) we can see that for all of these top 20 expressed genes the highest expression in found in the gonads (specifically, three ovaries and two testes), which is different from the other two _Stigmatopora_ spp. and all of the _Syngnathus_ species that had the highest expression in the liver. There are also a few genes that appear to be highly expressed across all of the tissue types. 

With this heatmap we can also pull out the names of the Trinity genes that are showing this high expression and BLAST the corresponding sequences to see what the genes are.

```{r heatmap_genes}
#Pull out the corresponding trinity_geneIDS that are plotted in the heatmap
heatmap_TG <- cbind(row.names(assay(vsd)[select,]))
#Run once to create the file and then commented out
#write.table(heatmap_TG,
 #           'heatmap_trinitygenes.txt',
  #          sep = "",
   #         quote=FALSE,
    #        row.names = FALSE,
     #       col.names = FALSE)
```

### Sample-dist Heatmap

We can then generate a sample-to-sample distances heatmap that will give an overview of the similarities and differences between our samples. The darker the colour, the more similar the samples are. The diagonal line of very dark blue represents the comparisson between the same samples.

```{r sample-dist, echo=FALSE, fig.cap="Sample-to-sample distances heatmap showing the similarities and differences between all of the samples. The darker the colour, the more similar the samples are. The diagonal line of very dark blue represents the comparisons between the same samples.", warning=FALSE}
#Calculate the sample-to-sample distances from the transformed count matrix
sampleDists <- dist(t(assay(vsd)))

#Transform into a matrix
sampleDistsMatrix <- as.matrix(sampleDists)
rownames(sampleDistsMatrix) <- paste(vsd$sex, vsd$organ, sep = "-")
colnames(sampleDistsMatrix) <- vsd$ID

#Run the heatmap
colors <- colorRampPalette(rev(brewer.pal(9, 'Blues')))(255)
pheatmap(sampleDistsMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists, 
         col = colors)
```

This is a definite improvement on the original sample distribution heatmap, although we still have one female gill (S39) clustering with livers, one male gonad clustering with gills (S46), and one female liver clustering with gonads (S32), so we might need to sort that out. We'll see what the PCA plots look like first. 

### PCA plots

```{r pca-pairs-plot, echo=FALSE, fig.cap="Principal components analysis pairsplot showing PCA axis 1 - 4."}
#Generate the PCA data to create the pairs plot, using the PCAtools package
p <- pca(assay(vsd), 
         metadata = colData(dds_SA))

#Create the pairs plot
pairsplot(p,
          components = getComponents(p, c(1:4)),
          colby = "organ",
          triangle = FALSE,
          hline = 0, vline = 0,
          colkey = c("gill" = "#6E8B3D75", 
                     "gonad" = "#EEB42275", 
                     "liver" = "#EE826275"),
          shape = "sex",
          shapekey = c(16, 17),
          pointSize = 3,
          margingaps = unit(c(0.1, 0.1, 0.1, 0.1), 'cm'),
          legendPosition = "left", legendIconSize = 3, legendLabSize = 10)

```

Huzzah! The organs are all clustering with their organ groups.

Next we can look at PCA plots for our samples to see how our different samples are grouping.
```{r pca-plot, echo=FALSE, message=FALSE, warning=FALSE}

#Generate the PCA dataframe
pca <- prcomp(t(assay(vsd)))
pca_plotting_data<-as.data.frame(pca$x)

#Add the metadata to the PCA dataset
pca_plotting_data$organ <- samples$organ
pca_plotting_data$sex <- samples$sex
pca_plotting_data$ID <- samples$ID

#Calculate the percent variation attributed to each axis 
percentVar <- round(p$variance, digits = 2)


my_colors <- c("gill" = "#6E8B3D75", "gonad" = "#EEB42275", "liver" = "#EE826275")
my_shapes <- c(16, 17)

ggplot(pca_plotting_data, aes(x=PC1, y=PC2, color = organ, shape = sex)) +
  scale_color_manual(values = my_colors) +
  scale_shape_manual(values = my_shapes) +
  geom_point(size = 5) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) +
  geom_hline(yintercept = 0,
             linetype = 'dashed') +
  geom_vline(xintercept = 0,
             linetype = 'dashed') + 
  geom_text(aes(label = ID), vjust = -1, hjust = 0.5)
```