---
title: "Genetic diversity and Intersexual Fst in _Syngnathus floridae_"
author: "Coley Tosto"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    fig_caption: yes
    keep_tex: yes
    number_sections: no
    toc: false
header-includes: >
  \usepackage{lipsum}
  \usepackage{float}
  \floatplacement{figure}{H}
editor_options:
  chunk_output_type: console
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir='../',fig_path="../imgs/")
```

``` {r library, message = FALSE, warning = FALSE}
#This is a cohesive list of all the libraries used in this document
library(DESeq2)
```

``` {r functions, include = FALSE}
```

``` {r read-data, message = FALSE, warning = FALSE}
#Tajima's D data calculated with ANGSD
tajD <- read.delim("data/floridae.thetas.idx.pestPG", header = TRUE)

#Intersexual Fst calculated with ANGSD
fst <- read.delim("data/floridae_fm_fst.txt", header = FALSE)

#The abundance matrix generated via salmon and tximport to be used for the DE analysis
txi.salmon <- readRDS("data/txi.salmon.floride.RDS")

#The samples file generated for tximport to be used for DE analysis
samples <- read.table("FL_samples.txt", header = TRUE)

#Make sure the conditions are in the samples file as a factor
samples$Sex <- as.factor(samples$Sex)
samples$Organ <- as.factor(samples$Organ)
```

# Generating BAM files
`Bowtie2` was used to map the processed reads back to the reference transcriptome, generating SAM files that will be converted to BAM files and coordinate sorted with `SAMtools` and then used by ANGSD to calculate measurements of Tajima's $D$ and intersexual $F_{st}$.

`Bowtie2` was installed inside of a conda environment on the RCC named `trinity`. Bowtie2 v2.5.0 was used. `SAMtools` was installed in the same conda environment, v1.18 was used. The following script was then used to generate the index, map the reads back to it to create the SAM files, then convert those SAM files to BAM files, and finally coordinate sort the BAM files.

```{bash, eval = FALSE, file="bash/bowtie2_alignment.sh"}

```

The script was run as `nohup bash bash_scripts/bowtie2_alignment.sh trinity_supertran_floridae.fasta bowtie2_index/floridae floridae_kmer_corrected/ floridae_SAM/ floridae_BAM/ > bt2.log 2>&1 &`.

# Running ANGSD
ANGSD will be used for the calculation of both Tajima's $D$ and also Intersexual $F_{st}$. It was installed in the `shared/` folder on the RCC following the instructions given on their [website](http://www.popgen.dk/angsd/index.php/Installation) as:

```
wget http://popgen.dk/software/download/angsd/angsd0.940.tar.gz
tar xf angsd0.940.tar.gz

cd htslib;make;cd ..

cd angsd
make HTSSRC=../htslib
cd ..
```

ANGSD version 0.940-dirty was used for the analysis.

The following script was then used to filter the BAM files, estimate site frequency spectrum (SFS), calculate the thetas and the calculate Tajima's $D$ when `$ALLSFS=true and $THETAS=true`. It was then also used to filter BAM files and estimate SFS **separately** for males and females, combine them together and then calculate intersexual $F_{st}$ when `$INDSFS=true and $MFFST=true`.

```{bash, eval=FALSE, file='bash/angsd.sh'}


```

#### When $ALLSFS=true and $THETAS=true

Following the [ANGSD website](angsd/index.php/Thetas,Tajima,Neutrality_tests) we can see the first step for calculate Tajima's $D$ is to **filter and estimate SFS**:

  - To get the estimates of SFS you first generate a `.saf` file (site allele frequency likelihood) followed by an optimization of the `.saf` file which will then estimate the site frequency spectrum.
    - `doSaf 1` was used to calculate saf based on individual genotype likelihoods assuming HWE.
    - `GL 1` was used since SAMtools was used to generate the BAM files.
    - Because we don't have the ancestral state, we estimated the folded SFS by giving `-anc` the reference transcriptome and applying `-fold 1` to `realSFS`.
    
  - For filtering, many options were used:
    - `minMapQ`: set the minimum mapping quality (20 was used here)
    - `remove_bads`: removes reads with a flag above 255 (set to 1 for remove)
    - `uniqueOnly`: when set to 1, removes reads that have multiple best hits
    - `only_proper_pairs`: when set to 1, includes only pairs of reads where both mates mapped correctly
    - `minQ`: Minimum base quality score (set to 13 here)
    - `minInd`: Remove if there was data in less than X individuals (4 here).
    
The second step is to then **calculate the thetas for each site**:
  - This is done using the `.sfa.idx` and the `.sfs` files from the step before
  
Lastly, with the output from the thetas calculation (`.thetas.idx`) we can **estimate Tajima's $D$**.

#### When $INDSFS=true and $MFFST=true

Following the [ANGSD website](http://popgen.dk/angsd/index.php/Fst) we can see the first step for calculating $F_{st}$ is similar to Tajima's $D$, we have to **filter and estimate SFS**. The difference here is that we calculate the SAF **SEPARATELY** for the different populations and then put them together with the `realSFS` function. Because we want to look at differences between males and females our different "populations" are the two sexes.

  - `-fold 1` was given to `realSFS` again because we do not have the ancestral state.
  
  - The same filtering that was used for Tajima's $D$ was applied here, however this time the RNA-seq data was filtered to only included bases where we had data in more than half of the individuals for male and females separately.
    - This was done by setting `minInd` to 7
    
After getting the male-female SFSs and combining them, intersexual $F_{st}$ can be calculated with `realSFS fst`.

# Tajima's D
After running ANGSD we can start to look through some of the results. To start with I am plotting the site frequency spectrum that was generated by ANGSD.

```{r SFS, echo=FALSE, fig.cap="Site frequency spectrum calculated by ANGSD."}
##Read in the data
sfs <- scan('data/floridaeFolded.sfs')

#The variable categories of the sfs
sfs <- sfs[-c(1,length(sfs))]
sfs <- sfs/sum(sfs)

#Plot it
barplot(sfs, names=1:length(sfs),
        main='SFS', ylab = "Proportions",
        col = 'indianred4')

```

I then want to look at the overall distribution of Tajima's $D$ that was calculated. Additionally, I want to see if each row of the Tajima's $D$ dataset corresponds to an individual gene. There are `r nrow(tajD)` rows in the Tajima's $D$ dataset. If each row was one gene than I would have originally expected to see `r nrow(txi.salmon$counts)` rows as that is how many Trinity genes we have in our assembly. If we look as the unique IDs for our "chromosome" column in the Tajima's $D$ dataset we can see that there are `r length(unique(tajD$Chr))` unique IDs. This does tell us that each row is likely to be an individual Trinity gene and it is possible that the low amount of rows here compared to the number of genes in out transcriptome could be due to the filtering restrictions we applied above.

```{r TajHist, echo=FALSE, fig.cap="Histogram showing the ditribution of Tajima's D values."}
hist(tajD$Tajima,
     xlab = "Tajima's D",
     ylab = "Number of genes",
     main = "")
```

It appears that the low number of genes that we are seeing here may be due to a lack of sufficient individuals being aligned to the same regions, and then whether those regions are variable enough for the pop gen statistics to be calculated. I am going to try aligning my reads with a different software and see if that changes anything.

# Re-generating SAM/BAM files with BWA
`Bowtie2` was originally used to map the processed reads back to the reference transcriptome, but now I am going to try `Burrows-Wheeler Aligner` (`BWA`) to generate the SAM files. Converting the SAM files to BAM files nd coordinate sorting them will still be done with `SAMtools` v1.18 as I did before.

`BWA` was installed inside of a conda environment on the RCC named `bwa`. `BWA` v0.7.17 was used. `SAMtools` was installed in the same conda environment. The following script was then used to generate the index, map the reads back to it to create the SAM files, then convert those SAM files to BAM files, and finally coordinate sort the BAM files.

```{bash, eval = FALSE, file="bash/bwa_alignment.sh"}

```

The script was run as `nohup bash bash_scripts/bwa_alignment.sh trinity_supertran_floridae.fasta bwa_index/floridae floridae_kmer_corrected/ floridae_SAM_bwa/ floridae_BAM_bwa/ > bwa.log 2>&1 &`.